---
title: "Seurat: the standard scRNA-seq workflow"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figs/01_seurat_intro-"
    comment: "#>"
---

This tutorial walks through the standard Seurat workflow on a dataset of 2,700
peripheral blood mononuclear cells (PBMCs) sequenced on the 10x Genomics
platform. By the end you will have gone from a raw count matrix to annotated
cell types on a UMAP.

Everything here runs inside the course container. If you have not built it yet,
go back to the in-class assignment before starting.

## Setup

Set a working directory on `/xdisk`. Replace the NetID if you are working
somewhere else.

```{r}
#| label: setup
library(Seurat)
library(dplyr)
library(patchwork)
library(ggplot2)

WORK <- file.path("/xdisk/darrenc/cmm_523", Sys.info()[["user"]], "seurat_intro")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)
```

## Get the data

The PBMC dataset is distributed by 10x Genomics as a gzipped tarball. This only
downloads if the file is not already present, so it is safe to re-run.

```{r}
#| label: download
tarball <- file.path(WORK, "pbmc3k_filtered_gene_bc_matrices.tar.gz")

if (!file.exists(tarball)) {
  download.file(
    "https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz",
    tarball, mode = "wb"
  )
  untar(tarball, exdir = WORK)
}

list.files(file.path(WORK, "filtered_gene_bc_matrices", "hg19"))
```

Three files: `barcodes.tsv`, `genes.tsv`, and `matrix.mtx`. This is the standard
Cell Ranger output format — a sparse matrix plus the row and column names for
it.

## Build a Seurat object

`Read10X()` reads those three files into a sparse matrix of counts, with genes
as rows and cells as columns.

```{r}
#| label: read10x
pbmc.data <- Read10X(data.dir = file.path(WORK, "filtered_gene_bc_matrices", "hg19"))

pbmc <- CreateSeuratObject(
  counts = pbmc.data,
  project = "pbmc3k",
  min.cells = 3,
  min.features = 200
)

pbmc
```

The two filters do different things. `min.cells = 3` drops genes detected in
fewer than three cells — these carry almost no information and slow everything
down. `min.features = 200` drops cells with fewer than 200 genes detected, which
are usually empty droplets or dying cells.

## Quality control

The standard QC metrics are the number of genes per cell, the number of UMIs per
cell, and the percentage of reads coming from mitochondrial genes. High
mitochondrial content usually means a dying or lysed cell, whose cytoplasmic RNA
has leaked out.

`PercentageFeatureSet()` computes the percentage of counts belonging to a set of
genes. In human data mitochondrial genes are prefixed `MT-`.

```{r}
#| label: qc-metrics
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

head(pbmc@meta.data, 5)
```

```{r}
#| label: qc-violin
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
#| label: qc-scatter
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

The scatter plots are worth reading carefully. `nCount_RNA` against
`nFeature_RNA` should be tightly correlated — cells with more UMIs detect more
genes. Points falling below that line are suspicious.

Now filter. These thresholds are specific to this dataset; they are not
universal constants.

```{r}
#| label: qc-filter
pbmc <- subset(
  pbmc,
  subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5
)

pbmc
```

## Normalization

Cells differ in sequencing depth, so raw counts are not comparable between them.
The default normalization divides each count by the total counts for that cell,
multiplies by 10,000, and log-transforms the result.

```{r}
#| label: normalize
pbmc <- NormalizeData(pbmc)
```

## Variable features

Most genes are not informative for distinguishing cell types. Restricting to the
genes that vary most across cells sharpens the downstream structure and cuts the
computation substantially.

```{r}
#| label: variable-features
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(pbmc), 10)
top10
```

```{r}
#| label: variable-feature-plot
plot1 <- VariableFeaturePlot(pbmc)
LabelPoints(plot = plot1, points = top10, repel = TRUE)
```

Notice what comes out on top: PPBP (platelets), LYZ (monocytes), GNLY and GNG11.
These are marker genes for specific populations, which is exactly what you would
hope a variance-based selection would find.

## Scaling

PCA requires that genes be centered and scaled, otherwise highly expressed genes
dominate purely because of their magnitude.

```{r}
#| label: scale
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

## Linear dimensional reduction

```{r}
#| label: pca
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))

print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
#| label: pca-loadings
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```

```{r}
#| label: pca-dimplot
DimPlot(pbmc, reduction = "pca") + NoLegend()
```

`DimHeatmap()` is a useful way to judge how many components carry real signal.
Each panel shows the cells and genes with the most extreme loadings on that
component, ordered. A component with clear block structure is capturing
something; one that looks like noise probably is.

```{r}
#| label: dimheatmap-1
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```

```{r}
#| label: dimheatmap-many
#| fig-height: 14
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```

The elbow plot ranks components by the variance they explain. The "elbow" — the
point where the curve flattens — is a rough guide to how many to keep.

```{r}
#| label: elbow
ElbowPlot(pbmc)
```

There is an elbow around PC 9–10 here. We will use 10. This choice is a judgment
call, and it is worth checking that your conclusions do not depend heavily on it.

## Clustering

Seurat clusters cells by building a shared nearest neighbor graph and then
partitioning it. The `resolution` parameter controls granularity: higher values
give more clusters.

```{r}
#| label: cluster
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)

head(Idents(pbmc), 5)
```

```{r}
#| label: umap
pbmc <- RunUMAP(pbmc, dims = 1:10)
DimPlot(pbmc, reduction = "umap")
```

## Finding marker genes

A cluster is only useful if you can say what it is. `FindAllMarkers()` tests
each cluster against all the others and returns the genes that distinguish it.

```{r}
#| label: markers
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)

pbmc.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 3) %>%
  ungroup()
```

You can also test specific comparisons. Here, cluster 5 against clusters 0 and 3:

```{r}
#| label: markers-specific
cluster5.markers <- FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3))
head(cluster5.markers, n = 5)
```

Plotting known markers across the UMAP is the quickest way to orient yourself.

```{r}
#| label: feature-plot
#| fig-height: 10
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A",
                               "FCGR3A", "LYZ", "PPBP", "CD8A"))
```

```{r}
#| label: marker-heatmap
#| fig-height: 10
pbmc.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 10) %>%
  ungroup() -> top10_markers

DoHeatmap(pbmc, features = top10_markers$gene) + NoLegend()
```

## Assigning cell type identities

With canonical markers in hand, the clusters can be named. This mapping is
specific to this dataset at this resolution — if your clustering differs, so
will the numbering.

| Cluster | Markers | Cell type |
|---|---|---|
| 0 | IL7R, CCR7 | Naive CD4 T |
| 1 | CD14, LYZ | CD14+ Mono |
| 2 | IL7R, S100A4 | Memory CD4 T |
| 3 | MS4A1 | B |
| 4 | CD8A | CD8 T |
| 5 | FCGR3A, MS4A7 | FCGR3A+ Mono |
| 6 | GNLY, NKG7 | NK |
| 7 | FCER1A, CST3 | DC |
| 8 | PPBP | Platelet |

```{r}
#| label: annotate
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B",
                     "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet")

stopifnot(length(new.cluster.ids) == length(levels(pbmc)))
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)

DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

## Save your work

```{r}
#| label: save
plot <- DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 4.5) +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(axis.title = element_text(size = 18),
        legend.text = element_text(size = 18)) +
  guides(colour = guide_legend(override.aes = list(size = 10)))

ggsave(file.path(WORK, "output", "pbmc3k_umap.jpg"),
       plot = plot, height = 7, width = 12)

saveRDS(pbmc, file = file.path(WORK, "output", "pbmc3k_final.rds"))
```

Keep that `.rds`. Several later tutorials start from a processed object, and
re-running this one takes long enough to be annoying.

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the [Seurat PBMC3K guided
tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
(Satija Lab), updated for Seurat 5 and R 4.6.*
