---
title: "SingleR: automated cell type annotation"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figs/02_singler-"
    comment: "#>"
---

In the [Seurat intro](01_seurat_intro.md) you annotated clusters by hand: look
up canonical markers, decide what each cluster is, type in the names. That works,
but it is slow, it requires you to already know the markers, and it is hard to
defend when someone asks why cluster 4 is a CD8 T cell and not something else.

SingleR takes a different approach. Given a reference dataset of known cell
types, it correlates each of your cells against that reference and assigns the
best match. It annotates **cells**, not clusters, which means you can also use it
to check whether your clustering is carving the data at sensible joints.

## Setup

```{r}
#| label: setup
library(Seurat)
library(SingleR)
library(celldex)
library(scRNAseq)
library(pheatmap)
library(viridis)
library(ggplot2)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

if (nzchar(Sys.getenv("CMM523_NETID"))) NETID <- Sys.getenv("CMM523_NETID")

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "singler")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

# Rendering only: keep the multi-GB knitr cache off /home.
knitr::opts_chunk$set(
  cache.path = file.path(
    Sys.getenv("CMM523_CACHE", unset = "/xdisk/darrenc/darrenc/cmm523_cache"),
    "02_singler/"
  )
)

WORK
```

## Get a reference

`celldex` provides several curated reference datasets. The Human Primary Cell
Atlas is a good general-purpose starting point for blood and immune data.

The first call downloads and caches the reference, which takes a minute. The
cache location is set by the container, so it lands on `/xdisk` rather than in
your home directory.

```{r}
#| label: reference
ref <- celldex::HumanPrimaryCellAtlasData()
ref
```

Each reference has two label columns. `label.main` is coarse (T cell, B cell,
monocyte); `label.fine` is granular (naive CD4 T, memory CD8 T, and so on).
Which you want depends on the question.

```{r}
#| label: reference-labels
table(ref$label.main)[1:15]
```

The Human Primary Cell Atlas covers many tissues, most of which are irrelevant
for PBMCs. Subsetting to the relevant lineages makes the assignment both faster
and more accurate — a reference full of hepatocytes gives SingleR opportunities
to be wrong in uninteresting ways.

```{r}
#| label: subset-reference
keep <- grepl("DC|B_cell|T_cells|Monocyte|NK_cell|^HSC_CD34", ref$label.main)
ref <- ref[, keep]

table(ref$label.main)
```

## Get some data to annotate

We will use a PBMC dataset from `scRNAseq` rather than the object from the last
tutorial, so this tutorial stands alone.

```{r}
#| label: test-data
sce <- scRNAseq::KotliarovPBMCData()
sce
```

SingleR expects log-normalized expression. The dataset arrives as a
`SingleCellExperiment`, but since the rest of this course works in Seurat, we
will convert it and normalize the Seurat way. SingleR does not care which
object the matrix came from.

```{r}
#| label: to-seurat
pbmc <- CreateSeuratObject(counts = counts(sce), project = "kotliarov")
pbmc <- NormalizeData(pbmc)

pbmc
```

## Run SingleR

```{r}
#| label: singler-main
norm_counts <- LayerData(pbmc, assay = "RNA", layer = "data")

pred.main <- SingleR(
  test = norm_counts,
  ref = ref,
  labels = ref$label.main,
  de.method = "wilcox"
)

table(pred.main$pruned.labels, useNA = "ifany")
```

The `de.method = "wilcox"` argument tells SingleR to pick marker genes by
Wilcoxon test rather than the default t-test. This is the recommended setting
for single-cell references.

`pruned.labels` is the important column. SingleR assigns every cell a label, but
then prunes assignments where the score is not meaningfully better than the
alternatives. Those become `NA`. A cell with `NA` is one SingleR is not confident
about, which is more useful than a confident wrong answer.

## Assess the assignments

The score heatmap shows, for each cell, how well it matched every label in the
reference. Good assignments look like a single bright band.

```{r}
#| label: score-heatmap
#| fig-height: 7
plotScoreHeatmap(pred.main)
```

The delta distribution is the gap between the best score and the median score
for each cell. Small deltas mean the cell matched everything about equally well,
which is a warning sign.

```{r}
#| label: delta-plot
#| fig-height: 7
plotDeltaDistribution(pred.main, ncol = 4, dots.on.top = FALSE)
```

## Fine-grained labels

```{r}
#| label: singler-fine
pred.fine <- SingleR(
  test = norm_counts,
  ref = ref,
  labels = ref$label.fine,
  de.method = "wilcox"
)

head(sort(table(pred.fine$pruned.labels, useNA = "ifany"), decreasing = TRUE), 15)
```

Fine labels are more informative and less reliable. Distinguishing naive from
central memory CD4 T cells from transcriptome alone is genuinely hard, and the
pruning is correspondingly more aggressive. Use fine labels when you need the
resolution, and check the delta plot before you trust them.

## Comparing SingleR to your clusters

This is where SingleR earns its keep. If you have already clustered your data,
cross-tabulating clusters against SingleR labels tells you whether the clusters
correspond to recognizable cell types.

```{r}
#| label: crosstab
pbmc <- AddMetaData(pbmc, pred.main$pruned.labels, col.name = "SingleR_main")

# A quick standard clustering to compare against
pbmc <- FindVariableFeatures(pbmc, nfeatures = 2000)
pbmc <- ScaleData(pbmc)
pbmc <- RunPCA(pbmc, npcs = 20)
pbmc <- FindNeighbors(pbmc, dims = 1:20)
pbmc <- FindClusters(pbmc, resolution = 0.5)

tab <- table(cluster = pbmc$seurat_clusters, label = pbmc$SingleR_main)
tab
```

```{r}
#| label: crosstab-heatmap
#| fig-height: 7
pheatmap(log10(tab + 10), color = viridis::viridis(100))
```

Read this heatmap carefully, because it is diagnostic in both directions. A
cluster that maps cleanly onto one label is well behaved. A cluster spread
across several labels may be under-clustered. Several clusters mapping to the
same label may mean you over-clustered, or may mean there is real substructure
the reference cannot name.

The single most useful summary is the dominant label per cluster:

```{r}
#| label: dominant-label
apply(tab, 1, function(x) names(which.max(x)))
```

## Labels on a UMAP

```{r}
#| label: umap-labels
#| fig-height: 7
pbmc <- RunUMAP(pbmc, dims = 1:20)

p1 <- DimPlot(pbmc, group.by = "seurat_clusters", label = TRUE) +
  NoLegend() + ggtitle("Seurat clusters")
p2 <- DimPlot(pbmc, group.by = "SingleR_main", label = TRUE, label.size = 3) +
  NoLegend() + ggtitle("SingleR labels")

p1 + p2
```

One detail worth noting in the code above: we pulled the expression matrix with
`LayerData(pbmc, assay = "RNA", layer = "data")`. If you find older code using
`GetAssayData(slot = "data")`, that is the Seurat 4 spelling — in Seurat 5 the
term is "layer" rather than "slot", and the old argument is deprecated.

## A caution about references

SingleR is only as good as its reference, and this is not a minor caveat. If
your cells are a type the reference does not contain, SingleR will still assign
them a label — the closest available match, confidently or not. Tumor cells
annotated against a healthy-tissue reference are a classic way to get
nonsense that looks authoritative.

Always check the delta distribution, always look at what fraction of cells got
pruned to `NA`, and treat a label as a hypothesis rather than a fact.

## Save your work

```{r}
#| label: save
saveRDS(pred.main, file = file.path(WORK, "output", "singler_main.rds"))
saveRDS(pred.fine, file = file.path(WORK, "output", "singler_fine.rds"))
```

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the [SingleR book](https://bioconductor.org/books/release/SingleRBook/)
and the biostatsquid SingleR tutorial, updated for Bioconductor 3.23 and R 4.6.*
