---
title: "Integration and label transfer"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figs/03_integration-"
    comment: "#>"
---

Almost no real study is a single sample. You have cells from several donors, or
several timepoints, or several technologies, and you want to analyze them
together. The problem is that if you simply concatenate them and cluster, the
dominant structure is usually the batch rather than the biology — cells separate
by which experiment they came from, not by what kind of cell they are.

Integration is the family of methods that addresses this. This tutorial covers
the Seurat approach and one of its most useful applications: transferring
annotations from a labeled reference onto unlabeled query cells.

## Before you start

This tutorial needs about **24 GB of memory** and runs in **30–40 minutes**.

```bash
interactive -a cusanovichlab -n 8 -t 03:00:00 --mem=32G
```

Integration holds several copies of the data in memory at once, so it is more
demanding than the single-sample workflow.

## Setup

```{r}
#| label: setup
#| cache: false
library(Seurat)
library(SeuratData)
library(patchwork)
library(ggplot2)
library(dplyr)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

if (nzchar(Sys.getenv("CMM523_NETID"))) NETID <- Sys.getenv("CMM523_NETID")

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "integration")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

# SeuratData datasets live in a shared read-only library on /xdisk. They are
# data, not code, so they sit outside the container -- baking a 200 MB dataset
# into every image would be wasteful, and the container is meant to define your
# software environment, not carry your data. If a dataset is missing from the
# shared library, the fallback below installs it into your own space.
SHARED_DATA <- "/xdisk/darrenc/cmm_523/references/Rdatalib"
MY_DATA     <- file.path("/xdisk/darrenc/cmm_523", NETID, "Rdatalib")
dir.create(MY_DATA, recursive = TRUE, showWarnings = FALSE)
.libPaths(c(MY_DATA, SHARED_DATA, .libPaths()))

knitr::opts_chunk$set(
  cache.path = file.path(
    Sys.getenv("CMM523_CACHE", unset = "/xdisk/darrenc/darrenc/cmm523_cache"),
    "03_integration/"
  )
)

options(future.globals.maxSize = 8000 * 1024^2)
WORK
```

## The data

`SeuratData` distributes curated datasets for exactly this purpose. `panc8` is
human pancreatic islet cells from eight donors across four different
single-cell technologies — a harder integration problem than most real studies,
which makes it a good demonstration.

```{r}
#| label: install-data
# panc8 should already be present in the shared library. The InstallData()
# call is only a fallback -- it needs somewhere writable, which is why the
# setup chunk above put your own directory on the library path first.
if (!requireNamespace("panc8.SeuratData", quietly = TRUE)) {
  message("panc8 not found in the shared library; downloading (117 MB)...")
  InstallData("panc8")
}
data("panc8", package = "panc8.SeuratData")

# panc8 was serialized under Seurat v3 and predates several slots that
# Seurat 5 objects have. Loading it as-is fails with errors like
# "no slot of name 'images'". UpdateSeuratObject() migrates it forward.
# Any Seurat object you are handed from a paper, a collaborator, or an older
# SeuratData package may need this -- objects do not migrate themselves.
panc8 <- UpdateSeuratObject(panc8)

panc8
```

```{r}
#| label: inspect
table(panc8$tech)
table(panc8$celltype)[1:10]
```

Four technologies. `celltype` is the published annotation, which we will use
both to check the integration and, later, as the labels to transfer.

## Why integration is needed

Before integrating, look at what happens if you don't. Seurat 5 keeps samples as
separate **layers** within one object, so splitting is a metadata operation
rather than a restructuring.

```{r}
#| label: split
panc8[["RNA"]] <- split(panc8[["RNA"]], f = panc8$tech)
panc8
```

```{r}
#| label: unintegrated
panc8 <- NormalizeData(panc8)
panc8 <- FindVariableFeatures(panc8)
panc8 <- ScaleData(panc8)
panc8 <- RunPCA(panc8)

panc8 <- FindNeighbors(panc8, dims = 1:30, reduction = "pca")
panc8 <- FindClusters(panc8, resolution = 0.5, cluster.name = "unintegrated_clusters")
panc8 <- RunUMAP(panc8, dims = 1:30, reduction = "pca",
                 reduction.name = "umap.unintegrated")
```

```{r}
#| label: unintegrated-plot
#| fig-width: 12
DimPlot(panc8, reduction = "umap.unintegrated",
        group.by = c("tech", "celltype"), label = FALSE) +
  plot_layout(guides = "collect")
```

This is the problem, drawn. Cells of the same type sit in different places
depending on which technology produced them. Any clustering here recovers
technology, not biology.

## Integration

Seurat 5 handles this with a single function, `IntegrateLayers()`. You pass a
method; the function returns a new dimensional reduction in which shared cell
types are co-embedded across batches.

```{r}
#| label: integrate-cca
panc8 <- IntegrateLayers(
  object = panc8,
  method = CCAIntegration,
  orig.reduction = "pca",
  new.reduction = "integrated.cca",
  verbose = FALSE
)
```

Note what did *not* happen: no `FindIntegrationAnchors()`, no `IntegrateData()`,
no separate "integrated" assay. If you find older tutorials using that pattern,
they are written for Seurat v4. The result is comparable; the interface is
simpler.

Now cluster and embed using the integrated reduction rather than the raw PCA.

```{r}
#| label: cluster-integrated
panc8 <- FindNeighbors(panc8, reduction = "integrated.cca", dims = 1:30)
panc8 <- FindClusters(panc8, resolution = 0.5, cluster.name = "cca_clusters")
panc8 <- RunUMAP(panc8, reduction = "integrated.cca", dims = 1:30,
                 reduction.name = "umap.cca")
```

```{r}
#| label: integrated-plot
#| fig-width: 12
DimPlot(panc8, reduction = "umap.cca",
        group.by = c("tech", "celltype"), label = FALSE) +
  plot_layout(guides = "collect")
```

Compare against the unintegrated version above. Technologies now overlap; cell
types separate. That is what success looks like.

## Rejoining layers

While the layers are split, expression-level operations act per-layer. Before
running differential expression or anything else that treats the data as one
dataset, rejoin them.

```{r}
#| label: rejoin
panc8 <- JoinLayers(panc8)
panc8
```

Forgetting this is a common source of confusing errors in Seurat 5 — functions
either fail or silently operate on only one batch.

## Judging integration

Integration is a balance between two failure modes, and both are easy to reach.
Under-correct and batch effects remain. Over-correct and you erase real
biological differences — including, in the worst case, merging distinct cell
types because they came from different samples.

There is no single number that tells you which you have. The practical checks
are: do batches mix, do known cell types stay separate, and do marker genes
still behave as they should?

```{r}
#| label: check-mixing
table(panc8$cca_clusters, panc8$tech)
```

Rows where one technology dominates are suspicious — either genuine biology
specific to one experiment, or incomplete integration.

```{r}
#| label: check-markers
#| fig-height: 8
FeaturePlot(panc8, reduction = "umap.cca",
            features = c("INS", "GCG", "SST", "PPY"))
```

Insulin, glucagon, somatostatin and pancreatic polypeptide mark beta, alpha,
delta and gamma cells. Each should light up one region. If two of these overlap
after integration, you have over-corrected.

## Label transfer

Now the practical payoff. You have a well-annotated reference and a new
unlabeled dataset. Rather than annotating from scratch, project the reference
labels onto the query.

Split the pancreas data: three technologies as reference, one held out as query.

```{r}
#| label: split-ref-query
pancreas.ref   <- subset(panc8, tech %in% c("celseq2", "smartseq2", "fluidigmc1"))
pancreas.query <- subset(panc8, tech == "celseq")

pancreas.ref   <- NormalizeData(pancreas.ref)
pancreas.ref   <- FindVariableFeatures(pancreas.ref)
pancreas.ref   <- ScaleData(pancreas.ref)
pancreas.ref   <- RunPCA(pancreas.ref)
pancreas.ref   <- RunUMAP(pancreas.ref, dims = 1:30, return.model = TRUE)

pancreas.query <- NormalizeData(pancreas.query)
```

`return.model = TRUE` matters. Without it the UMAP cannot be reused to project
new cells, and the mapping step below will fail.

```{r}
#| label: find-anchors
anchors <- FindTransferAnchors(
  reference = pancreas.ref,
  query = pancreas.query,
  dims = 1:30,
  reference.reduction = "pca"
)
```

Anchors are pairs of cells — one in the reference, one in the query — that
appear to be the same cell type. They are the bridge across which labels travel.

```{r}
#| label: map-query
pancreas.query <- MapQuery(
  anchorset = anchors,
  reference = pancreas.ref,
  query = pancreas.query,
  refdata = list(celltype = "celltype"),
  reference.reduction = "pca",
  reduction.model = "umap"
)
```

```{r}
#| label: mapping-plot
#| fig-width: 12
p1 <- DimPlot(pancreas.ref, reduction = "umap", group.by = "celltype",
              label = TRUE, label.size = 3, repel = TRUE) +
  NoLegend() + ggtitle("Reference")

p2 <- DimPlot(pancreas.query, reduction = "ref.umap",
              group.by = "predicted.celltype",
              label = TRUE, label.size = 3, repel = TRUE) +
  NoLegend() + ggtitle("Query, projected")

p1 + p2
```

## How good are the predictions?

Because these cells came with published annotations, we can grade the transfer
directly — a luxury you will not have on real data.

```{r}
#| label: accuracy
table(predicted = pancreas.query$predicted.celltype,
      actual = pancreas.query$celltype)

mean(pancreas.query$predicted.celltype == pancreas.query$celltype)
```

On real data the substitute is the prediction score, which reports how confident
the transfer was for each cell.

```{r}
#| label: prediction-scores
VlnPlot(pancreas.query, features = "predicted.celltype.score",
        group.by = "predicted.celltype", pt.size = 0) +
  NoLegend() +
  labs(title = "Prediction confidence by assigned label")
```

Read this the way you read the SingleR delta plot. A label assigned with a low
score is a label to distrust. Cell types poorly represented in the reference
tend to show up here as broad, low distributions — which is the honest signal
that the reference could not really speak to those cells.

## Save your work

```{r}
#| label: save
saveRDS(panc8, file = file.path(WORK, "output", "panc8_integrated.rds"))
saveRDS(pancreas.query, file = file.path(WORK, "output", "pancreas_query_mapped.rds"))
```

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the Seurat [introduction to
integration](https://satijalab.org/seurat/articles/integration_introduction.html)
and [mapping and
annotation](https://satijalab.org/seurat/articles/integration_mapping.html)
vignettes, updated to the Seurat 5 `IntegrateLayers` interface.*
