---
title: "Harmony: comparing integration methods"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figs/04_harmony-"
    cache.path: "/xdisk/darrenc/darrenc/cmm523_cache/04_harmony/"
    comment: "#>"
---

In the [previous tutorial](03_integration.md) you integrated the pancreas data
with CCA. That is one method among several, and the choice matters: methods
differ in how aggressively they correct, how they scale to large datasets, and
what they assume about your batches.

Harmony is the most widely used alternative. It works in PCA space, iteratively
nudging cells toward batch-mixed cluster centroids. It is fast, it scales well,
and it corrects more aggressively than CCA — which is sometimes what you want
and sometimes exactly what you don't.

This tutorial runs both on the same data so you can see the difference rather
than take anyone's word for it.

> **About this tutorial.** This is our own version of a published tutorial,
> rewritten to run in the course container — see the credits at the bottom for
> the original. The original is worth reading too. It is what you will find when
> you search for this analysis, and comparing the two is good practice for the
> thing you will do constantly in your own work: taking a tutorial written for
> someone else's setup and making it run on yours.

## Before you start

We rendered this tutorial with:

```bash
interactive -a cusanovichlab -n 8 -t 03:00:00
```

`interactive` allocates memory per core — 4 GB each by default — so that is **32
GB** in total. Note there is no `--mem` flag: memory comes from the number of
cores you ask for. This one runs two integration methods on the same data.

It took about **8 minutes** to run when we did it. Ask for more time than you
expect to need — a job that hits its limit is killed part way through.

When R runs out of memory on the cluster, the scheduler kills it with no error
message — the session simply stops mid-command. If that ever happens to you,
here or anywhere else, memory is the first thing to check.

## Setup

```{r}
#| label: setup
#| cache: false
library(Seurat)
library(SeuratData)
library(harmony)
library(patchwork)
library(ggplot2)
library(dplyr)

# CHANGE THIS to your NetID.
NETID <- "your_netid"


WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "harmony")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

# SeuratData datasets live in a shared read-only library on /xdisk. They are
# data, not code, so they sit outside the container -- baking a 200 MB dataset
# into every image would be wasteful, and the container is meant to define your
# software environment, not carry your data. If a dataset is missing from the
# shared library, the fallback below installs it into your own space.
SHARED_DATA <- "/xdisk/darrenc/cmm_523/references/Rdatalib"
MY_DATA     <- file.path("/xdisk/darrenc/cmm_523", NETID, "Rdatalib")
dir.create(MY_DATA, recursive = TRUE, showWarnings = FALSE)
.libPaths(c(MY_DATA, SHARED_DATA, .libPaths()))


options(future.globals.maxSize = 8000 * 1024^2)
WORK
```

## Data and baseline processing

Same dataset as the integration tutorial, so the comparison is fair.

```{r}
#| label: data
if (!requireNamespace("panc8.SeuratData", quietly = TRUE)) {
  message("panc8 not found in the shared library; downloading (117 MB)...")
  InstallData("panc8")
}
data("panc8", package = "panc8.SeuratData")

# panc8 was serialized under Seurat v3 and predates several slots that
# Seurat 5 objects have. Loading it as-is fails with errors like
# "no slot of name 'images'". UpdateSeuratObject() migrates it forward.
# Any Seurat object you are handed from a paper, a collaborator, or an older
# SeuratData package may need this -- objects do not migrate themselves.
panc8 <- UpdateSeuratObject(panc8)

panc8[["RNA"]] <- split(panc8[["RNA"]], f = panc8$tech)

panc8 <- NormalizeData(panc8)
panc8 <- FindVariableFeatures(panc8)
panc8 <- ScaleData(panc8)
panc8 <- RunPCA(panc8)

panc8
```

## Running Harmony

In Seurat 5, Harmony is available through the same `IntegrateLayers()` interface
as CCA. Only the `method` argument changes.

```{r}
#| label: harmony
panc8 <- IntegrateLayers(
  object = panc8,
  method = HarmonyIntegration,
  orig.reduction = "pca",
  new.reduction = "harmony",
  verbose = FALSE
)
```

That interchangeability is worth pausing on. Integration methods are not
fundamentally different kinds of operation — they all take a dimensional
reduction contaminated by batch and return one that is less so. Swapping one
argument is the whole difference.

You will also encounter `RunHarmony()` in the wild, which is harmony's own
interface and still perfectly current:

```{r}
#| label: runharmony-alt
#| eval: false
panc8 <- RunHarmony(panc8, group.by.vars = "tech")
```

Both work. `IntegrateLayers()` is preferable here only because it makes the
comparison below symmetric.

## Running CCA for comparison

```{r}
#| label: cca
panc8 <- IntegrateLayers(
  object = panc8,
  method = CCAIntegration,
  orig.reduction = "pca",
  new.reduction = "integrated.cca",
  verbose = FALSE
)
```

## Three embeddings, one object

An advantage of the Seurat 5 design: multiple reductions coexist in the same
object, so nothing has to be recomputed to compare them.

```{r}
#| label: embeddings
panc8 <- RunUMAP(panc8, dims = 1:30, reduction = "pca",
                 reduction.name = "umap.unintegrated")
panc8 <- RunUMAP(panc8, dims = 1:30, reduction = "harmony",
                 reduction.name = "umap.harmony")
panc8 <- RunUMAP(panc8, dims = 1:30, reduction = "integrated.cca",
                 reduction.name = "umap.cca")

Reductions(panc8)
```

```{r}
#| label: compare-tech
#| fig-width: 15
#| fig-height: 5
p1 <- DimPlot(panc8, reduction = "umap.unintegrated", group.by = "tech") +
  ggtitle("No integration") + NoLegend()
p2 <- DimPlot(panc8, reduction = "umap.harmony", group.by = "tech") +
  ggtitle("Harmony") + NoLegend()
p3 <- DimPlot(panc8, reduction = "umap.cca", group.by = "tech") +
  ggtitle("CCA")

p1 + p2 + p3
```

```{r}
#| label: compare-celltype
#| fig-width: 15
#| fig-height: 5
p1 <- DimPlot(panc8, reduction = "umap.unintegrated", group.by = "celltype") +
  ggtitle("No integration") + NoLegend()
p2 <- DimPlot(panc8, reduction = "umap.harmony", group.by = "celltype") +
  ggtitle("Harmony") + NoLegend()
p3 <- DimPlot(panc8, reduction = "umap.cca", group.by = "celltype") +
  ggtitle("CCA")

p1 + p2 + p3
```

Look at both figures together. The first asks whether batches mixed; the second
asks whether cell types stayed distinct. A method that wins the first and loses
the second has not helped you.

## Quantifying batch mixing

Eyeballing UMAPs is a poor way to decide anything — UMAP distances are not
meaningful and the layout depends on the seed. A crude but honest numeric check
is to ask, for each cell, what fraction of its nearest neighbors come from a
different batch. Well-mixed data gives a high fraction.

```{r}
#| label: mixing-metric
neighbor_purity <- function(obj, reduction, batch_var = "tech", k = 30) {
  emb <- Embeddings(obj, reduction)[, 1:30]
  nn <- RANN::nn2(emb, k = k + 1)$nn.idx[, -1]
  batch <- obj[[batch_var]][, 1]
  mean(apply(nn, 1, function(idx) mean(batch[idx] != batch[1])))
}

data.frame(
  reduction = c("pca", "harmony", "integrated.cca"),
  cross_batch_neighbors = c(
    neighbor_purity(panc8, "pca"),
    neighbor_purity(panc8, "harmony"),
    neighbor_purity(panc8, "integrated.cca")
  )
)
```

Higher is more mixed. Treat this as a rough diagnostic rather than a score to
optimize — pushing it to 1.0 would mean the batches are indistinguishable, which
is also what happens when you have destroyed the biology.

## Clustering on each reduction

```{r}
#| label: cluster-both
panc8 <- FindNeighbors(panc8, reduction = "harmony", dims = 1:30)
panc8 <- FindClusters(panc8, resolution = 0.5, cluster.name = "harmony_clusters")

panc8 <- FindNeighbors(panc8, reduction = "integrated.cca", dims = 1:30)
panc8 <- FindClusters(panc8, resolution = 0.5, cluster.name = "cca_clusters")

agreement <- table(harmony = panc8$harmony_clusters, cca = panc8$cca_clusters)
agreement
```

The table is hard to read at a glance, so plot it. Scaling each row to sum to
one shows, for each Harmony cluster, how its cells were split among the CCA
clusters.

```{r}
#| label: cluster-agreement-heatmap
#| fig-height: 7
row_frac <- sweep(agreement, 1, rowSums(agreement), "/")
pheatmap::pheatmap(
  row_frac,
  color = viridis::viridis(100),
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = FALSE,
  main = "Fraction of each Harmony cluster in each CCA cluster"
)
```

If the two methods agree, each row has one bright square: every Harmony cluster
maps onto a single CCA cluster. Do not expect those squares to fall on a
diagonal. Each method numbers its clusters independently, so cluster 3 in one has
no reason to be cluster 3 in the other.

Rows where the colour is spread across several squares are where the methods
disagree about which cells belong together — and those cells are worth looking
at, because they are where the method choice actually changes your conclusions.

```{r}
#| label: rejoin
panc8 <- JoinLayers(panc8)
```

## Which should you use?

There is no general answer, which is itself the point worth taking away.

Harmony is fast and scales to very large datasets. It corrects aggressively,
which helps with strong batch effects and hurts when your batches differ
biologically — different treatments, different timepoints, different disease
states. If a real biological difference happens to align with your batch
structure, Harmony will remove it and the result will look clean.

CCA is slower and more conservative. It tends to preserve batch-associated
biology better, at the cost of leaving more technical variation behind.

The honest workflow is the one this tutorial demonstrates: run more than one,
compare, and check that your conclusions do not depend on which you picked. If
they do, that is a finding about the fragility of your result, not a reason to
pick the method that gives the answer you prefer.

## Save your work

```{r}
#| label: save
saveRDS(panc8, file = file.path(WORK, "output", "panc8_harmony_cca.rds"))
```

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the [Harmony
vignettes](https://portals.broadinstitute.org/harmony/) and the Seurat 5
integration documentation.*
