---
title: "Pseudobulk differential expression with edgeR"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figs/05_pseudobulk-"
    cache.path: "/xdisk/darrenc/darrenc/cmm523_cache/05_pseudobulk/"
    comment: "#>"
---

By now you have found marker genes with `FindAllMarkers()`, which tests every
cell against every other cell. That works for identifying cell types, but it is
the wrong tool for asking whether a gene differs between *conditions*.

The reason is that cells from the same donor are not independent observations.
If you sequence 5,000 cells from one patient, you do not have 5,000 independent
measurements of that patient's biology — you have one patient, measured 5,000
times. Treating those cells as independent inflates your sample size enormously
and produces p-values that are far too small. Run a cell-level test between two
conditions with three donors each and you will get thousands of "significant"
genes, most of which are telling you about donor identity rather than
condition.

Pseudobulk analysis fixes this by summing counts across all cells of a given
type within each sample, producing one profile per sample per cell type. You
then analyze those profiles with the same tools people have used for bulk
RNA-seq for fifteen years, which are well-tested and which treat samples as the
unit of replication — because they are.

> **About this tutorial.** This is our own version of a published tutorial,
> rewritten to run in the course container — see the credits at the bottom for
> the original. The original is worth reading too. It is what you will find when
> you search for this analysis, and comparing the two is good practice for the
> thing you will do constantly in your own work: taking a tutorial written for
> someone else's setup and making it run on yours.

## Before you start

We rendered this tutorial with:

```bash
interactive -a cusanovichlab -n 4 -t 02:00:00
```

`interactive` allocates memory per core — 4 GB each by default — so that is **16
GB** in total. Note there is no `--mem` flag: memory comes from the number of
cores you ask for. Pseudobulk profiles are small: the heavy lifting collapses
ten thousand cells into a few dozen columns.

It took about **2 minutes** to run when we did it. Ask for more time than you
expect to need — a job that hits its limit is killed part way through.

When R runs out of memory on the cluster, the scheduler kills it with no error
message — the session simply stops mid-command. If that ever happens to you,
here or anywhere else, memory is the first thing to check.

## Setup

```{r}
#| label: setup
#| cache: false
library(Seurat)
library(edgeR)
library(ggplot2)
library(pheatmap)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "pseudobulk")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

SHARED <- "/groups/darrenc/cmm_523/references/edgeR_pal"

WORK
```

## The data

We will use the same dataset as the pseudobulk case study in the edgeR User's
Guide (section 4.10), so you can compare your results directly against theirs —
a useful check that nothing has gone wrong.

The data come from a single-cell atlas of human breast tissue (Pal et al. 2021,
*EMBO Journal*). This subset contains normal breast epithelium from 13 donors,
already clustered by the original authors, and trimmed to 2,000 genes to keep it
small.

```{r}
#| label: data
seu_file <- file.path(SHARED, "SeuratObj.rds")

if (!file.exists(seu_file)) {
  seu_file <- file.path(WORK, "SeuratObj.rds")
  if (!file.exists(seu_file)) {
    download.file("https://bioinf.wehi.edu.au/edgeR/UserGuideData/SeuratObj.rds",
                  seu_file, mode = "wb")
  }
}

raw <- readRDS(seu_file)

# This object was saved years ago, and it was subset to 2,000 genes without its
# gene-level metadata being updated to match. So it carries a meta.features
# table describing the full gene set, and a var.features list naming genes that
# are no longer in it. Seurat does not complain on load, but the first operation
# that revalidates the assay fails with "'meta.features' must have the same
# number of rows as 'data'".
#
# Rather than patch that, rebuild a clean object from the counts and the cell
# metadata, which is all we actually need. This is a useful move to know: when
# an object you were given misbehaves, you can usually extract the parts that
# matter and start again.
# Note `layer`, not `slot`. The `slot` argument is defunct in SeuratObject 5 --
# not merely deprecated, so it errors rather than warns. Older code and
# tutorials you find will use `slot`; this is the Seurat 5 spelling.
counts <- SeuratObject::GetAssayData(raw, assay = "RNA", layer = "counts")
meta   <- raw@meta.data

seu <- CreateSeuratObject(counts = counts, meta.data = meta)
seu
```

```{r}
#| label: inspect
head(seu@meta.data)
table(seu$group)
table(seu$seurat_clusters)
```

`group` identifies the donor each cell came from, and `seurat_clusters` holds the
cell clusters from the original analysis. Those are the two things pseudobulk
needs: a unit of replication and the groups you want to compare.

## A first look at the cells

The edgeR guide shows these clusters on a t-SNE plot. We will use UMAP, which you
have already seen.

```{r}
#| label: umap
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- RunUMAP(seu, dims = 1:20)

DimPlot(seu, reduction = "umap", group.by = "seurat_clusters", label = TRUE) +
  NoLegend()
```

The clusters come from the original authors, not from us. Our UMAP is a fresh
embedding of the same cells, so it will not look identical to their t-SNE — but
the same cluster labels should still group together.

```{r}
#| label: umap-donor
DimPlot(seu, reduction = "umap", group.by = "group") +
  ggtitle("Cells coloured by donor")
```

Compare the two plots. If cells separated by donor rather than by cluster, that
would be a warning that donor differences were as large as cell type differences.
Here the donors are mixed within each cluster, which is what you want to see.

## Building pseudobulk profiles

edgeR provides `Seurat2PB()`, which sums the counts for every combination of
sample and cluster.

```{r}
#| label: pseudobulk
y <- Seurat2PB(seu, sample = "group", cluster = "seurat_clusters")

# Seurat2PB copies whatever gene-level metadata the Seurat object was carrying
# into y$genes, and edgeR prints that alongside its results. Our object picked
# up columns from FindVariableFeatures() when we made the UMAP above, which
# would push logFC and the p-values off the right-hand edge of every results
# table. Keep just the gene names.
y$genes <- data.frame(gene = rownames(y), row.names = rownames(y))

y
head(y$samples)
```

What came back is a `DGEList` — edgeR's container for bulk RNA-seq. Columns are
now donor-by-cluster combinations rather than cells: about ten thousand columns
became fewer than a hundred.

## Filtering

First, drop pseudobulk profiles that are too thin to be worth anything. A profile
summed from a handful of cells is mostly noise. `Seurat2PB()` does not report how
many cells went into each profile, so library size is the stand-in.

```{r}
#| label: filter-samples
summary(y$samples$lib.size)

keep.samples <- y$samples$lib.size > 5e4
table(keep.samples)

# Fail loudly rather than silently emptying the object.
stopifnot(sum(keep.samples) >= 3)

y <- y[, keep.samples]
```

Then drop genes not expressed at a useful level. `filterByExpr()` takes the group
structure into account rather than applying a flat cutoff.

```{r}
#| label: filter-genes
keep.genes <- filterByExpr(y, group = y$samples$cluster,
                           min.count = 10, min.total.count = 20)
table(keep.genes)

y <- y[keep.genes, , keep = FALSE]
dim(y)
```

## Normalization

```{r}
#| label: normalize
y <- normLibSizes(y)
summary(y$samples$norm.factors)
```

`normLibSizes()` computes TMM normalization factors, correcting for the fact that
a handful of very highly expressed genes can otherwise make everything else look
depleted. You may see older code call this `calcNormFactors()` — same thing,
previous name.

## Looking at the structure

`plotMDS()` gives a view of the data analogous to UMAP, where the distance
between each pair of points characterizes the similarity of those pseudobulk
samples. We expect to see clustering of samples from the same cell type dominate
over clustering of samples from the same subject.

```{r}
#| label: mds
#| fig-height: 7
cluster <- as.factor(y$samples$cluster)
plotMDS(y, pch = 16, col = c(2:8)[cluster], main = "MDS")
legend("bottomright", legend = paste0("cluster ", levels(cluster)),
       pch = 16, col = 2:8, cex = 0.8)
```

## The design

We want to find genes that distinguish each cluster from the others, while
accounting for differences between donors.

```{r}
#| label: design
donor <- factor(y$samples$sample)
design <- model.matrix(~ cluster + donor)
colnames(design) <- gsub("donor", "", colnames(design))
colnames(design)[1] <- "Int"
head(design)
```

Read the formula rather than the code. `~ cluster + donor` fits a baseline for
each donor and estimates the cluster effects on top of that. A gene that simply
differs between people is absorbed by the donor terms, and does not contaminate
the comparison between cell types. That is what a cell-level test cannot do.

## Dispersion

```{r}
#| label: dispersion
y <- estimateDisp(y, design, robust = TRUE)
plotBCV(y)
```

The biological coefficient of variation is roughly the typical relative
variability of a gene between replicates.

```{r}
#| label: qlfit
fit <- glmQLFit(y, design, robust = TRUE)
plotQLDisp(fit)
```

## Marker genes for each cluster

To find markers, we compare each cluster against the average of all the others.
That comparison has to be written as a contrast.

```{r}
#| label: contrasts
ncls <- nlevels(cluster)
contr <- rbind(matrix(1 / (1 - ncls), ncls, ncls),
               matrix(0, ncol(design) - ncls, ncls))
diag(contr) <- 1
contr[1, ] <- 0
rownames(contr) <- colnames(design)
colnames(contr) <- paste0("cluster", levels(cluster))
contr
```

Each column is one test. The `1` on the diagonal picks out one cluster; the
negative fractions average over the rest; the donor rows are zero because donor
is a nuisance we have adjusted for, not something we are testing.

```{r}
#| label: test
qlf <- list()
for (i in 1:ncls) {
  qlf[[i]] <- glmQLFTest(fit, contrast = contr[, i])
  qlf[[i]]$comparison <- paste0("cluster", levels(cluster)[i], "_vs_others")
}

topTags(qlf[[1]], n = 10L)
```

```{r}
#| label: summary
dt <- lapply(lapply(qlf, decideTests), summary)
do.call("cbind", dt)
```

That table is worth comparing against the edgeR guide. If your counts of up- and
down-regulated genes per cluster are close to theirs, the analysis is working as
intended.

## Heatmap of top markers

```{r}
#| label: heatmap
#| fig-height: 10
top <- 20
topMarkers <- list()
for (i in 1:ncls) {
  ord <- order(qlf[[i]]$table$PValue, decreasing = FALSE)
  up  <- qlf[[i]]$table$logFC > 0
  topMarkers[[i]] <- rownames(y)[ord[up][1:top]]
}
topMarkers <- unique(unlist(topMarkers))

lcpm  <- cpm(y, log = TRUE)
annot <- data.frame(cluster = paste0("cluster ", cluster))
rownames(annot) <- colnames(y)

pheatmap(lcpm[topMarkers, ],
         breaks = seq(-2, 2, length.out = 101),
         color = colorRampPalette(c("blue", "white", "red"))(100),
         scale = "row",
         cluster_cols = TRUE, border_color = NA,
         fontsize_row = 5,
         treeheight_row = 70, treeheight_col = 70,
         cutree_cols = ncls,
         clustering_method = "ward.D2",
         show_colnames = FALSE,
         annotation_col = annot)
```

Each column is a pseudobulk sample and each row a marker gene. If the markers
are doing their job, samples from the same cluster sit together regardless of
which donor they came from, and each cluster has its own block of highly
expressed genes.

## A note on conditions

This tutorial compares cell types. The same machinery is what you would use to
compare *conditions* — treated versus untreated, disease versus healthy — and
that is where pseudobulk matters most. Comparing two obviously different cell
types, the choice between a cell-level test and a pseudobulk test mostly changes
the length of your gene list. Comparing two conditions with a handful of donors
each, it changes whether your result is real.

## Save your work

```{r}
#| label: save
saveRDS(y,   file = file.path(WORK, "output", "pseudobulk_dgelist.rds"))
saveRDS(qlf, file = file.path(WORK, "output", "pseudobulk_markers.rds"))
```

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from section 4.10 of the [edgeR User's
Guide](https://bioconductor.org/packages/release/bioc/html/edgeR.html), using the
same data, updated for current edgeR (`normLibSizes`, native `Seurat2PB`) and
Seurat 5.*
