---
title: "Pseudobulk differential expression with edgeR"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figs/05_pseudobulk-"
    comment: "#>"
---

By now you have found marker genes with `FindAllMarkers()`, which tests every
cell against every other cell. That works for identifying cell types, and it is
the wrong tool for asking whether a gene differs between *conditions*.

The reason is that cells from the same donor are not independent observations.
If you sequence 5,000 cells from one patient, you do not have 5,000 independent
measurements of that patient's biology — you have one patient, measured 5,000
times. Treating those cells as independent inflates your sample size enormously
and produces p-values that are far too small. Run a cell-level test between two
conditions with three donors each and you will get thousands of "significant"
genes, most of which are telling you about donor identity rather than
condition.

Pseudobulk analysis fixes this by summing counts across all cells of a given
type within each sample, producing one profile per sample per cell type. You
then analyze those profiles with the same tools people have used for bulk
RNA-seq for fifteen years, which are well-tested and which treat samples as the
unit of replication — because they are.

## Before you start

This tutorial needs about **16 GB of memory** and runs in **15–20 minutes**.

```bash
interactive -a cusanovichlab -n 8 -t 02:00:00
```

The `interactive` command takes `-m` as memory **per core**, so `-n 8` with the
default 4 GB per core gives you 32 GB in total. There is no `--mem` flag.

## Setup

```{r}
#| label: setup
#| cache: false
library(Seurat)
library(SeuratData)
library(edgeR)
library(ggplot2)
library(dplyr)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

if (nzchar(Sys.getenv("CMM523_NETID"))) NETID <- Sys.getenv("CMM523_NETID")

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "pseudobulk")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

SHARED_DATA <- "/groups/darrenc/cmm_523/references/Rdatalib"
MY_DATA     <- file.path("/xdisk/darrenc/cmm_523", NETID, "Rdatalib")
dir.create(MY_DATA, recursive = TRUE, showWarnings = FALSE)
.libPaths(c(MY_DATA, SHARED_DATA, .libPaths()))

knitr::opts_chunk$set(
  cache.path = file.path(
    Sys.getenv("CMM523_CACHE", unset = "/xdisk/darrenc/darrenc/cmm523_cache"),
    "05_pseudobulk/"
  )
)

WORK
```

## The data

We will reuse the pancreas dataset from the [integration
tutorial](03_integration.md). It is well suited to this: eight donors, five
technologies, and published cell type labels. Donor is our unit of replication.

```{r}
#| label: data
if (!requireNamespace("panc8.SeuratData", quietly = TRUE)) {
  InstallData("panc8")
}
data("panc8", package = "panc8.SeuratData")
panc8 <- UpdateSeuratObject(panc8)

table(panc8$dataset, panc8$celltype)[, 1:6]
```

Look at that table before going further, because it determines what questions
you can ask. Each row is a donor, each column a cell type, and the numbers are
cell counts. Zeros and single digits are cell types that donor barely
contributed — those will produce pseudobulk profiles you should not trust, and
we filter them out below.

## Building pseudobulk profiles

edgeR provides `Seurat2PB()`, which does the aggregation for you. You tell it
which metadata column identifies the sample and which identifies the cluster.

```{r}
#| label: pseudobulk
y <- Seurat2PB(panc8, sample = "dataset", cluster = "celltype")

y
```

What came back is a `DGEList` — edgeR's bulk RNA-seq container. Columns are now
sample-by-celltype combinations rather than cells. The object went from ~15,000
columns to a few dozen, which is the whole point.

```{r}
#| label: inspect-pb
head(y$samples, 10)
dim(y)
```

## Filtering

Two filters matter here, and they do different jobs.

First, drop pseudobulk profiles that are too thin to be worth anything. A
profile aggregated from four cells is noise dressed up as data.

`Seurat2PB()` does not report how many cells went into each profile, so we use
library size as the stand-in: a profile built from very few cells has very few
total counts. There is no universal threshold. Look at the distribution and
pick somewhere sensible for your data rather than copying a number.

```{r}
#| label: lib-sizes
summary(y$samples$lib.size)
hist(log10(y$samples$lib.size), breaks = 20,
     xlab = "log10(library size)", main = "Pseudobulk profile depth")
```

```{r}
#| label: filter-cells
MIN_LIB <- 5e4

keep.samples <- y$samples$lib.size >= MIN_LIB
table(keep.samples)

# Fail loudly rather than silently returning an empty object. A filter that
# removes everything is a mistake, not a result, and an empty DGEList produces
# confusing errors several steps later rather than here.
stopifnot(sum(keep.samples) >= 3)

y <- y[, keep.samples]
dim(y)
```

Second, drop genes that are not expressed at a useful level. `filterByExpr()`
does this using the design of the experiment rather than a flat cutoff, which
is why you pass it the group structure.

```{r}
#| label: filter-genes
keep.genes <- filterByExpr(y, group = y$samples$cluster)
table(keep.genes)

y <- y[keep.genes, , keep.lib.sizes = FALSE]
dim(y)
```

## Normalization

```{r}
#| label: normalize
y <- normLibSizes(y)
head(y$samples)
```

`normLibSizes()` computes TMM normalization factors, correcting for the fact
that a handful of very highly expressed genes can otherwise make everything
else look depleted. If you find older code calling `calcNormFactors()`, that is
the previous name for this function — it still works, but `normLibSizes()` is
the current one.

## Looking at the structure

Before testing anything, look at how the samples relate to each other. An MDS
plot is the bulk RNA-seq equivalent of a UMAP, and it will tell you immediately
whether cell type or donor is the dominant source of variation.

```{r}
#| label: mds
#| fig-height: 7
cluster <- factor(y$samples$cluster)
plotMDS(y, col = as.numeric(cluster), pch = 16)
legend("topright", legend = levels(cluster),
       col = seq_along(levels(cluster)), pch = 16, cex = 0.7)
```

You want to see profiles grouping by cell type rather than by donor. If donors
separated instead, that would be a warning: the differences between people
would be larger than the differences between cell types, and any comparison you
made would be confounded.

## Setting up the test

We will ask which genes distinguish alpha cells from beta cells, while
accounting for donor. That "while accounting for donor" is the reason to do any
of this — it is what a cell-level test cannot do.

```{r}
#| label: design
donor   <- factor(y$samples$sample)
cluster <- factor(y$samples$cluster)

# Put alpha first so it becomes the reference level. Without this, R picks the
# alphabetically first cell type (acinar), and every "cluster" coefficient
# would be a comparison against acinar rather than against alpha. That is a
# perfectly valid analysis -- it is just not the one we said we were doing,
# and nothing in the output would tell you.
cluster <- relevel(cluster, ref = "alpha")

design <- model.matrix(~ donor + cluster)
colnames(design) <- gsub("cluster", "", colnames(design))
colnames(design) <- gsub("donor", "", colnames(design))

dim(design)
```

Reading the formula matters more than the code. `~ donor + cluster` says: fit a
baseline for each donor, then estimate the cell type effect on top of that. Any
gene that simply differs between people is absorbed by the donor terms and does
not contaminate the cell type comparison.

## Estimating dispersion

```{r}
#| label: dispersion
y <- estimateDisp(y, design, robust = TRUE)
plotBCV(y)
```

The biological coefficient of variation is the square root of the dispersion —
roughly, the typical relative variability of a gene between replicates. For
pseudobulk data this is usually higher than for bulk RNA-seq, because you are
also absorbing variation in how many cells contributed to each profile.

```{r}
#| label: qlfit
fit <- glmQLFit(y, design, robust = TRUE)
plotQLDisp(fit)
```

## Testing

```{r}
#| label: test
# Confirm what we are actually testing before testing it.
grep("beta", colnames(design), value = TRUE)

res <- glmQLFTest(fit, coef = "beta")
topTags(res, n = 15)
```

The table is sorted by p-value, so the top of it is whatever is most confidently
different — which need not be the genes you expect. Check the positive control
explicitly instead of hoping it appears:

```{r}
#| label: positive-control
all_res <- topTags(res, n = Inf)$table
all_res[c("INS", "GCG"), c("logFC", "PValue", "FDR")]
```

`INS` (insulin) should be strongly positive: up in beta cells. `GCG` (glucagon)
should be strongly negative: it marks alpha cells, our reference. If those two
come out the wrong way round, the contrast is inverted and everything
downstream is backwards — which is exactly the kind of error that produces a
publishable-looking gene list pointing in the wrong direction.

```{r}
#| label: summary
summary(decideTests(res))
```

```{r}
#| label: md-plot
plotMD(res, main = "Beta vs alpha")
abline(h = c(-1, 1), col = "blue", lty = 2)
```

Points above the line are up in beta, below are down in alpha. The blue lines
mark two-fold change.

## Comparing against the cell-level test

This is the claim the whole tutorial rests on, so it is worth measuring rather
than asserting.

Making the comparison fair takes care. `FindMarkers()` filters genes before
testing — by default it skips anything below `logfc.threshold` or expressed in
too few cells — so out of the box it tests a much smaller set of genes than
edgeR does. Turn those filters off, or you are comparing two different gene
universes and the counts mean nothing.

```{r}
#| label: compare-celllevel
alpha_beta <- subset(panc8, celltype %in% c("alpha", "beta"))
Idents(alpha_beta) <- alpha_beta$celltype

cell_level <- FindMarkers(
  alpha_beta, ident.1 = "beta", ident.2 = "alpha",
  logfc.threshold = 0,   # default 0.1 -- drops genes before testing
  min.pct = 0            # default 0.01 -- same
)

data.frame(
  test = c("pseudobulk (edgeR)", "cell-level (Seurat)"),
  genes_tested = c(nrow(y), nrow(cell_level)),
  n_significant = c(
    sum(decideTests(res) != 0),
    sum(cell_level$p_val_adj < 0.05)
  )
)
```

Look at the two columns together, not just the second. The counts only mean
something relative to how many genes each test considered.

The cell-level test calls a higher proportion of what it tested — and note that
it also tested more genes, because edgeR's `filterByExpr()` had already removed
those too sparsely expressed to be informative, while `FindMarkers()` with the
filters off keeps everything.

The gap here is real but not dramatic, and that is worth being honest about.
Alpha and beta cells are genuinely very different, so both tests find a lot, and
neither answer is absurd.

The gap matters much more in the case this method exists for: comparing two
*conditions* — treated versus untreated, say, with three donors each. There the
cell-level test will hand you thousands of genes while you have six biological
replicates, and its p-values will be reporting how many cells you sequenced
rather than how many people you studied. Sequence twice as many cells from the
same three donors and the p-values shrink, though you have learned nothing new
about the biology.

That is the asymmetry to remember. Comparing two obviously different cell types,
the choice of test changes the length of your list. Comparing two conditions
across a handful of donors, it changes whether your result is real.

The practical rule: cell-level tests to characterize and name clusters,
pseudobulk to compare groups you intend to make claims about.

## Save your work

```{r}
#| label: save
saveRDS(y,   file = file.path(WORK, "output", "pseudobulk_dgelist.rds"))
saveRDS(res, file = file.path(WORK, "output", "pseudobulk_results.rds"))

write.csv(topTags(res, n = Inf)$table,
          file = file.path(WORK, "output", "beta_vs_alpha.csv"))
```

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from section 4.10 of the [edgeR User's
Guide](https://bioconductor.org/packages/release/bioc/html/edgeR.html), updated
for current edgeR (`normLibSizes`, native `Seurat2PB`) and Seurat 5.*
