---
title: "Trajectory inference with monocle3"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figs/06_trajectory-"
    comment: "#>"
---

Clustering assumes cells fall into discrete groups. For many tissues that is a
reasonable approximation — a B cell is a B cell. But differentiation is not
discrete. A haematopoietic stem cell becoming an erythrocyte passes through a
continuum of states, and forcing that continuum into clusters throws away the
thing you probably care about: the order.

Trajectory inference fits a path through the data and places each cell along
it. The resulting coordinate is usually called **pseudotime** — not time in
hours, but position along the inferred path.

Two warnings before you start, because this is the technique in this course
most likely to produce a confident wrong answer.

A trajectory algorithm will fit a trajectory to anything. Give it discrete cell
types with no developmental relationship and it will still draw a path through
them, and that path will look perfectly plausible. The algorithm has no way to
tell you that your data has no trajectory in it.

And pseudotime is not time. Cells at the same pseudotime are at similar
transcriptional states, which may or may not mean they are at the same point in
a real process.

## Before you start

This tutorial needs about **32 GB of memory** and runs in **30–45 minutes**.

```bash
interactive -a cusanovichlab -n 8 -t 03:00:00
```

## Setup

```{r}
#| label: setup
#| cache: false
library(Seurat)
library(SeuratData)
library(monocle3)
library(ggplot2)
library(dplyr)
library(patchwork)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

if (nzchar(Sys.getenv("CMM523_NETID"))) NETID <- Sys.getenv("CMM523_NETID")

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "trajectory")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

SHARED_DATA <- "/groups/darrenc/cmm_523/references/Rdatalib"
MY_DATA     <- file.path("/xdisk/darrenc/cmm_523", NETID, "Rdatalib")
dir.create(MY_DATA, recursive = TRUE, showWarnings = FALSE)
.libPaths(c(MY_DATA, SHARED_DATA, .libPaths()))

knitr::opts_chunk$set(
  cache.path = file.path(
    Sys.getenv("CMM523_CACHE", unset = "/xdisk/darrenc/darrenc/cmm523_cache"),
    "06_trajectory/"
  )
)

options(future.globals.maxSize = 8000 * 1024^2)
WORK
```

## The data

Bone marrow, where haematopoiesis is actively happening — so there is a real
continuum to find, not one we are imposing.

```{r}
#| label: data
if (!requireNamespace("hcabm40k.SeuratData", quietly = TRUE)) {
  InstallData("hcabm40k")
}
data("hcabm40k", package = "hcabm40k.SeuratData")
bm <- UpdateSeuratObject(hcabm40k)

bm
```

40,000 cells is more than we need and enough to make every step slow.
Subsample.

```{r}
#| label: subsample
set.seed(42)
bm <- bm[, sample(ncol(bm), 10000)]
dim(bm)
```

## Standard processing first

Trajectory inference operates on an embedding, so we need one. This is the
ordinary Seurat workflow.

```{r}
#| label: process
bm <- NormalizeData(bm)
bm <- FindVariableFeatures(bm, nfeatures = 2000)
bm <- ScaleData(bm)
bm <- RunPCA(bm, npcs = 50)
bm <- FindNeighbors(bm, dims = 1:30)
bm <- FindClusters(bm, resolution = 0.5)
bm <- RunUMAP(bm, dims = 1:30)

DimPlot(bm, label = TRUE) + NoLegend()
```

```{r}
#| label: markers-check
#| fig-height: 8
FeaturePlot(bm, features = c("CD34", "GATA1", "HBB", "MPO", "CD79A", "CD3E"))
```

Orient yourself with these before going further. `CD34` marks progenitors —
wherever it is highest is where the trajectory should start. `GATA1` and `HBB`
mark the erythroid branch, `MPO` myeloid, `CD79A` B lymphoid, `CD3E` T. If you
can see those arranged around a common origin, there is a trajectory to find.

## Converting to a cell_data_set

monocle3 uses its own object class. We convert with a small helper rather than
`SeuratWrappers::as.cell_data_set()`, because SeuratWrappers pulls in a large
tree of spatial-transcriptomics packages for that one function.

```{r}
#| label: convert
source("seurat_to_cds.R")

cds <- seurat_to_cds(bm, reductions = c(pca = "PCA", umap = "UMAP"))
cds
```

Read `seurat_to_cds.R` if you have not. It is about thirty lines: pull the
counts, build a table of cell metadata and a table of gene metadata, hand all
three to `new_cell_data_set()`, and copy the reductions across. A
`cell_data_set` is not a mysterious object, and knowing what a wrapper actually
does is worth the ten minutes.

## Clustering in monocle3

monocle3 needs its own clustering, because it uses **partitions** — groups of
cells it considers connected — to decide where one trajectory ends and another
begins.

```{r}
#| label: cluster-cds
cds <- cluster_cells(cds, reduction_method = "UMAP")

plot_cells(cds, color_cells_by = "partition", show_trajectory_graph = FALSE,
           group_label_size = 4)
```

Partitions matter more than they look. Cells in different partitions get
separate trajectories with separate roots. If monocle3 splits something you
believe is one continuous process, the trajectory will be wrong in a way that
is hard to see afterwards.

## Learning the graph

```{r}
#| label: learn-graph
cds <- learn_graph(cds, use_partition = TRUE)

plot_cells(cds,
           color_cells_by = "seurat_clusters",
           label_groups_by_cluster = FALSE,
           label_leaves = FALSE,
           label_branch_points = TRUE,
           graph_label_size = 3)
```

The black line is the principal graph — the skeleton monocle3 fits through the
data. Branch points are where it thinks a decision happens.

## Choosing a root

Here is the step where biology has to enter, and where the method cannot help
you. Pseudotime is a distance from a starting point, and **you choose the
starting point.** Choose a different root and every cell's pseudotime changes.

We pick the cell with the highest `CD34` expression, because progenitors are
where haematopoiesis starts.

```{r}
#| label: root
cd34 <- LayerData(bm, assay = "RNA", layer = "data")["CD34", ]
root_cell <- names(which.max(cd34))

cat("root cell:", root_cell, "\n")
cat("CD34 (normalized):", round(max(cd34), 2), "\n")

cds <- order_cells(cds, root_cells = root_cell)
```

In an interactive session `order_cells(cds)` opens a picker and you click the
root yourself. That does not work in a rendered document, and it does not work
in a batch job — hence choosing programmatically here. It is also more
defensible: "the cell with the highest CD34" is a statement someone can
disagree with, while "the one I clicked" is not.

```{r}
#| label: pseudotime-plot
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           label_branch_points = FALSE,
           graph_label_size = 3)
```

## Checking it against known biology

A trajectory is a hypothesis. Test it against genes whose behaviour you already
know.

```{r}
#| label: pseudotime-genes
#| fig-height: 8
bm$pseudotime <- pseudotime(cds)[colnames(bm)]

FeaturePlot(bm, features = "pseudotime") +
  scale_colour_viridis_c() +
  ggtitle("Pseudotime")
```

```{r}
#| label: gene-trends
#| fig-height: 5
genes <- c("CD34", "GATA1", "HBB")
df <- data.frame(
  pseudotime = bm$pseudotime,
  t(as.matrix(LayerData(bm, assay = "RNA", layer = "data")[genes, ]))
)
df <- df[is.finite(df$pseudotime), ]

df |>
  tidyr::pivot_longer(-pseudotime, names_to = "gene", values_to = "expr") |>
  ggplot(aes(pseudotime, expr)) +
  geom_smooth(method = "loess", formula = y ~ x) +
  facet_wrap(~ gene, scales = "free_y") +
  theme_bw() +
  labs(title = "Expression along pseudotime")
```

`CD34` should fall as pseudotime increases; `GATA1` should rise and then fall;
`HBB` should rise late and steeply. If instead `CD34` rises, your root is at the
wrong end — flip it and re-run. That is a normal thing to have happen, and
catching it is exactly why you check against known markers rather than trusting
the plot.

## Genes that change along the trajectory

```{r}
#| label: graph-test
graph_res <- graph_test(cds, neighbor_graph = "principal_graph", cores = 4)

graph_res |>
  filter(q_value < 0.01) |>
  arrange(desc(morans_I)) |>
  select(gene_short_name, morans_I, q_value) |>
  head(20)
```

Moran's I measures spatial autocorrelation on the graph: genes whose expression
varies smoothly along the trajectory rather than randomly. High values are
genes that track the process.

```{r}
#| label: top-trajectory-genes
#| fig-height: 8
top <- graph_res |>
  filter(q_value < 0.01) |>
  arrange(desc(morans_I)) |>
  head(4) |>
  pull(gene_short_name)

plot_cells(cds, genes = top, show_trajectory_graph = FALSE,
           label_cell_groups = FALSE, label_leaves = FALSE)
```

## What this cannot tell you

Worth stating plainly, because trajectory plots are persuasive out of
proportion to their evidence.

The trajectory is a fit to your embedding, and the embedding depends on your
choice of variable features, number of PCs, and UMAP parameters. Change those
and the trajectory can change. Pseudotime ordering is only as good as the root,
which you chose. And a branch in the graph is not proof of a fate decision —
it can equally reflect two cell types that happen to sit near each other.

None of that makes the method useless. It makes it a hypothesis generator. The
right follow-up to a trajectory is an experiment, not a p-value.

## Save your work

```{r}
#| label: save
saveRDS(cds, file = file.path(WORK, "output", "trajectory_cds.rds"))
saveRDS(bm,  file = file.path(WORK, "output", "trajectory_seurat.rds"))
write.csv(graph_res, file = file.path(WORK, "output", "graph_test_results.csv"))
```

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the [monocle3
documentation](https://cole-trapnell-lab.github.io/monocle3/docs/trajectories/)
and the Seurat-wrappers monocle3 vignette, updated for Seurat 5 and rewritten
to avoid the SeuratWrappers dependency.*
