---
title: "Single-cell ATAC-seq with Signac"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  # Caching is OFF for this tutorial. knitr stores cached objects in a
  # lazy-load database that cannot hold a variable larger than 2 GB, and a
  # ChromatinAssay over ~100,000 peaks exceeds that. Turning cache on here
  # fails with "long vectors not supported yet: connections.c". The cost is
  # that every render re-scans the fragment file.
  cache: false
knitr:
  opts_chunk:
    fig.path: "figs/07_signac-"
    comment: "#>"
---

Everything so far has been RNA. scATAC-seq measures something different:
which regions of the genome are open, and therefore which regulatory elements a
cell is using. Where RNA tells you what a cell is currently making, ATAC tells
you something about what it is equipped to make.

The data has a different shape too, and that shape drives every methodological
difference you are about to meet. An scRNA count matrix is genes by cells, maybe
20,000 rows, with counts in the tens or hundreds. An scATAC matrix is peaks by
cells — often 100,000+ rows — and the values are mostly 0, 1, or 2. There are
only two copies of any locus in a diploid cell, so "how open is this peak" is
nearly a binary question per cell.

Sparse, near-binary, high-dimensional data does not suit the tools built for
RNA. Signac handles it with a different normalization (TF-IDF, borrowed from
text search) and a different dimensional reduction (LSI rather than PCA).

## Before you start

This tutorial needs about **32 GB of memory** and runs in **45–60 minutes**.
It reads the fragment file several times, which is most of that time.

```bash
interactive -a cusanovichlab -n 8 -t 03:00:00
```

## Setup

```{r}
#| label: setup
#| cache: false
library(Signac)
library(Seurat)
library(EnsDb.Hsapiens.v86)
# GenomeInfoDb provides standardChromosomes(), seqlevels() and genome().
# Signac imports these internally but does not attach them for you.
library(GenomeInfoDb)
library(GenomicRanges)
library(ggplot2)
library(patchwork)
library(dplyr)

set.seed(1234)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

if (nzchar(Sys.getenv("CMM523_NETID"))) NETID <- Sys.getenv("CMM523_NETID")

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "signac")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

SHARED <- "/groups/darrenc/cmm_523/references/pbmc_multiome"

WORK
```

## The data

10x Genomics Multiome on about 10,000 PBMCs — both RNA and ATAC measured in the
same cells. This tutorial uses only the ATAC half; the [RNA+ATAC
tutorial](08_rna_atac.md) comes back for the rest, so the download serves twice.

Three files:

- the **feature-by-cell matrix**, containing both gene expression and peaks
- the **ATAC fragment file**, a record of every sequenced fragment and the cell
  it came from
- its **index**, which lets Signac read regions of the fragment file without
  loading all of it

The fragment file is the one with no RNA equivalent, and it is what makes the
coverage plots at the end of this tutorial possible. The peak matrix has already
thrown away everything outside a peak; the fragment file still has it.

```{r}
#| label: data
files <- c(
  "pbmc_granulocyte_sorted_10k_filtered_feature_bc_matrix.h5",
  "pbmc_granulocyte_sorted_10k_atac_fragments.tsv.gz",
  "pbmc_granulocyte_sorted_10k_atac_fragments.tsv.gz.tbi"
)

base_url <- paste0("https://cf.10xgenomics.com/samples/cell-arc/1.0.0/",
                   "pbmc_granulocyte_sorted_10k/")

get_file <- function(f) {
  shared <- file.path(SHARED, f)
  if (file.exists(shared)) return(shared)

  local <- file.path(WORK, f)
  if (!file.exists(local)) {
    message("Downloading ", f, " ...")
    download.file(paste0(base_url, f), local, mode = "wb")
  }
  local
}

counts_file <- get_file(files[1])
frag_file   <- get_file(files[2])
invisible(get_file(files[3]))   # index must sit beside the fragment file
```

```{r}
#| label: load
inputdata <- Read10X_h5(counts_file)
names(inputdata)

atac_counts <- inputdata$Peaks
dim(atac_counts)
atac_counts[1:5, 1:3]
```

Look at those values. Mostly zeros, with the occasional 1 or 2. That is what
near-binary means in practice, and it is why the RNA toolkit does not transfer.

## Gene annotation

Signac needs to know where genes are so it can label plots and compute gene
activity. We pull that from an Ensembl database package.

```{r}
#| label: annotation
# This dataset was mapped against GRCh38-2020-A, so we need an hg38 annotation.
# EnsDb.Hsapiens.v86 is hg38; EnsDb.Hsapiens.v75 is hg19. Getting this wrong is
# not a loud failure -- the code runs, TSS enrichment is computed against
# coordinates that do not correspond to your peaks, every cell scores near zero,
# and your QC filter then silently discards the entire dataset. Always check
# which build your data was aligned to before choosing an annotation.
annotation <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)

# Ensembl calls chromosomes "1", "2", "X". UCSC's hg38 calls them "chr1",
# "chr2", "chrX". Rename directly rather than using seqlevelsStyle() -- the
# style-conversion route depends on GenomeInfoDb's mapping tables and is a
# recurring source of the error "Annotation genome does not match genome of
# the object".
seqlevels(annotation) <- paste0("chr", seqlevels(annotation))
genome(annotation) <- "hg38"

annotation[1:3]
```

## Building the object

```{r}
#| label: chromatin-assay
# Keep only the standard chromosomes. Peaks on scaffolds and alt contigs are
# not useful here and cause problems downstream.
grange_counts <- StringToGRanges(rownames(atac_counts), sep = c(":", "-"))
grange_use    <- seqnames(grange_counts) %in% standardChromosomes(grange_counts)
atac_counts   <- atac_counts[as.vector(grange_use), ]

chrom_assay <- CreateChromatinAssay(
  counts = atac_counts,
  sep = c(":", "-"),
  genome = "hg38",
  fragments = frag_file,
  min.cells = 10,
  min.features = 200,
  annotation = annotation
)

pbmc <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "peaks"
)

pbmc
```

```{r}
#| label: granges
granges(pbmc)[1:5]
```

Each feature is a genomic interval rather than a gene name. That is the other
structural difference from RNA data, and it is why you can ask questions of ATAC
data — what is near this peak, do these two peaks overlap — that make no sense
for a gene count matrix.

## Quality control

ATAC has its own QC metrics, and they measure things that have no RNA analogue.

```{r}
#| label: qc-metrics
pbmc <- NucleosomeSignal(object = pbmc)

# fast = FALSE keeps the per-position signal around the TSS, not just the
# summary score. TSSPlot() needs it; the default (fast = TRUE) computes the
# score and throws the profile away, and TSSPlot() then fails with
# "Position enrichment matrix not present in assay".
pbmc <- TSSEnrichment(object = pbmc, fast = FALSE)

head(pbmc@meta.data[, c("nCount_peaks", "TSS.enrichment", "nucleosome_signal")], 5)
```

Note where these came from: both were computed from the fragment file, not read
out of a metadata table. Cell Ranger ARC does not ship the per-barcode QC csv
that the older ATAC pipeline did, and relying on a file's columns is fragile
anyway — a renamed column gives you `NA`, and `NA < 0.05` quietly removes every
cell in your dataset.

What each metric is telling you:

**Nucleosome signal** is the ratio of fragments longer than one nucleosome to
those shorter. Tn5 cuts accessible DNA, so a good library shows clear laddering
— sub-nucleosomal fragments, then mono-, then di-nucleosomal. A high ratio means
poor chromatin digestion.

**TSS enrichment** is the ratio of signal at transcription start sites to
flanking background. Promoters are open in essentially every cell, so this works
as close to a universal positive control: low TSS enrichment means the assay did
not work in that cell, however many reads it has.

```{r}
#| label: qc-fragment-histogram
pbmc$nucleosome_group <- ifelse(pbmc$nucleosome_signal > 2, "NS > 2", "NS < 2")
FragmentHistogram(object = pbmc, group.by = "nucleosome_group")
```

That plot is worth dwelling on. The good group shows the nucleosomal laddering
described above. The bad group does not — you can see the failure directly in
the fragment length distribution rather than inferring it from a summary
statistic.

```{r}
#| label: qc-tss
pbmc$high.tss <- ifelse(pbmc$TSS.enrichment > 2, "High", "Low")
TSSPlot(pbmc, group.by = "high.tss") + NoLegend()
```

```{r}
#| label: qc-violin
#| fig-width: 12
#| fig-height: 5
VlnPlot(
  object = pbmc,
  features = c("nCount_peaks", "TSS.enrichment", "nucleosome_signal"),
  pt.size = 0.1,
  ncol = 3
)
```

Look at the distributions before choosing thresholds. A cut is only sensible
relative to what it is being applied to.

```{r}
#| label: qc-distributions
qc <- pbmc@meta.data[, c("nCount_peaks", "TSS.enrichment", "nucleosome_signal")]
summary(qc)
cat("cells with any NA in a QC metric:", sum(!complete.cases(qc)), "\n")
```

```{r}
#| label: qc-filter
before <- ncol(pbmc)

keep <- pbmc$nCount_peaks      > 1000 &
        pbmc$nCount_peaks      < 100000 &
        pbmc$nucleosome_signal < 2 &
        pbmc$TSS.enrichment    > 1

# What each criterion removes on its own. If one is discarding almost
# everything, that is the threshold to question -- or the sign that the metric
# was not computed the way you assumed.
data.frame(
  criterion = c("counts > 1000", "counts < 100000",
                "nucleosome < 2", "TSS > 1"),
  n_passing = c(
    sum(pbmc$nCount_peaks      > 1000,   na.rm = TRUE),
    sum(pbmc$nCount_peaks      < 100000, na.rm = TRUE),
    sum(pbmc$nucleosome_signal < 2,      na.rm = TRUE),
    sum(pbmc$TSS.enrichment    > 1,      na.rm = TRUE)
  ),
  of_total = before
)

pbmc <- pbmc[, which(keep)]
cat("kept", ncol(pbmc), "of", before, "cells\n")

# Fail here rather than several steps downstream. A filter that removes almost
# everything is a mistake, not a result.
stopifnot(ncol(pbmc) > 500)

pbmc
```

## Normalization and dimensional reduction

Here is where ATAC diverges most sharply from RNA.

```{r}
#| label: lsi
pbmc <- RunTFIDF(pbmc)
pbmc <- FindTopFeatures(pbmc, min.cutoff = "q0")
pbmc <- RunSVD(pbmc)

# How many components did we actually get? RunSVD() asks for 50 by default but
# can return fewer, and asking for dimensions that do not exist fails later
# with "subscript out of bounds" -- an error that points at RunUMAP rather than
# at the real cause here.
n_lsi <- ncol(Embeddings(pbmc, "lsi"))
n_lsi
```

**TF-IDF** comes from document retrieval. Treat each cell as a document and each
peak as a word: a peak open in every cell is uninformative, like the word "the",
while a peak open in a few cells is highly informative. TF-IDF weights peaks by
how rare they are. This is a better fit than log-normalization, which assumes a
count distribution ATAC data does not have.

**SVD on the TF-IDF matrix** is latent semantic indexing — the same technique
used for finding documents by topic. Together, TF-IDF followed by SVD is what
people mean by LSI in the ATAC literature.

```{r}
#| label: depth-correlation
DepthCor(pbmc)
```

This plot exists because of a specific, well-known artefact: **the first LSI
component usually captures sequencing depth rather than biology.** If the
correlation at component 1 is strongly negative, exclude it — which is why
everything below uses `dims = 2:30` rather than `1:30`.

Nothing warns you about this. Include component 1 and your clusters will
partly reflect how deeply each cell was sequenced, and the UMAP will look
perfectly reasonable.

Everything below therefore starts at component 2.

```{r}
#| label: cluster
# Skip component 1 (depth), use what is available up to 30.
use_dims <- 2:min(30, n_lsi)
range(use_dims)

pbmc <- RunUMAP(object = pbmc, reduction = "lsi", dims = use_dims)
pbmc <- FindNeighbors(object = pbmc, reduction = "lsi", dims = use_dims)
pbmc <- FindClusters(object = pbmc, algorithm = 3, resolution = 1.2, verbose = FALSE)

DimPlot(object = pbmc, label = TRUE) + NoLegend()
```

## Gene activity

Clusters of peaks are hard to interpret directly. The usual move is to summarize
accessibility over each gene body and promoter, producing a matrix that looks
like gene expression and can be handled with familiar tools.

```{r}
#| label: gene-activity
gene.activities <- GeneActivity(pbmc)

pbmc[["RNA"]] <- CreateAssayObject(counts = gene.activities)
pbmc <- NormalizeData(
  object = pbmc,
  assay = "RNA",
  normalization.method = "LogNormalize",
  scale.factor = median(pbmc$nCount_RNA)
)

dim(gene.activities)
```

Be clear about what this is: a **proxy**, not a measurement. Accessibility over
a gene body correlates with expression, loosely. A gene can be open and not
transcribed. Gene activity is useful for recognizing cell types; it is not a
substitute for measuring RNA.

```{r}
#| label: gene-activity-plot
#| fig-height: 8
DefaultAssay(pbmc) <- "RNA"

FeaturePlot(
  object = pbmc,
  features = c("MS4A1", "CD3D", "LEF1", "NKG7", "TREM1", "LYZ"),
  pt.size = 0.1,
  max.cutoff = "q95",
  ncol = 3
)
```

Recognizable: `MS4A1` for B cells, `CD3D` for T cells, `LYZ` for monocytes. The
signal is noisier than the RNA equivalent, which is what you should expect from
a proxy.

## Coverage plots

This is the visualization with no RNA counterpart, and it is the reason the
fragment file was worth keeping.

```{r}
#| label: coverage
DefaultAssay(pbmc) <- "peaks"

CoveragePlot(
  object = pbmc,
  region = "MS4A1",
  extend.upstream = 5000,
  extend.downstream = 5000
)
```

Each row is a cluster, and the height is accessibility across the locus. You are
looking at the regulatory landscape of a gene, cell type by cell type: which
promoter is open, which enhancers are used, and by whom. A `FeaturePlot` tells
you a gene is on in B cells. This tells you *where* the chromatin is open to
make that happen.

```{r}
#| label: coverage-cd3
CoveragePlot(
  object = pbmc,
  region = "CD3D",
  extend.upstream = 5000,
  extend.downstream = 5000
)
```

## Save your work

```{r}
#| label: save
saveRDS(pbmc, file = file.path(WORK, "output", "pbmc_atac.rds"))
```

Keep this. The RNA+ATAC integration tutorial starts from it.

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the [Signac PBMC](https://stuartlab.org/signac/articles/pbmc_vignette)
and [multiomic](https://stuartlab.org/signac/articles/pbmc_multiomic) vignettes,
updated for Signac 1.17 and Seurat 5.*
