---
title: "Single-cell ATAC-seq with Signac"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  # Caching is OFF for this tutorial. knitr stores cached objects in a
  # lazy-load database that cannot hold a variable larger than 2 GB, and a
  # ChromatinAssay over ~100,000 peaks exceeds that. Turning cache on here
  # fails with "long vectors not supported yet: connections.c". The cost is
  # that every render re-scans the fragment file.
  cache: false
knitr:
  opts_chunk:
    fig.path: "figs/07_signac-"
    comment: "#>"
---

Everything so far has been RNA. scATAC-seq measures something different:
which regions of the genome are open, and therefore which regulatory elements a
cell is using. Where RNA tells you what a cell is currently making, ATAC tells
you something about what it is equipped to make.

The data has a different shape too, and that shape drives every methodological
difference you are about to meet. An scRNA count matrix is genes by cells, maybe
20,000 rows, with counts in the tens or hundreds. An scATAC matrix is peaks by
cells — often 100,000+ rows — and the values are mostly 0, 1, or 2. There are
only two copies of any locus in a diploid cell, so "how open is this peak" is
nearly a binary question per cell.

Sparse, near-binary, high-dimensional data does not suit the tools built for
RNA. Signac handles it with a different normalization (TF-IDF, borrowed from
text search) and a different dimensionality reduction (LSI rather than PCA).

> **About this tutorial.** This is our own version of a published tutorial,
> rewritten to run in the course container — see the credits at the bottom for
> the original. The original is worth reading too. It is what you will find when
> you search for this analysis, and comparing the two is good practice for the
> thing you will do constantly in your own work: taking a tutorial written for
> someone else's setup and making it run on yours.

## Before you start

We rendered this tutorial with:

```bash
interactive -a cusanovichlab -n 8 -t 03:00:00
```

`interactive` allocates memory per core — 4 GB each by default — so that is **32
GB** in total. Note there is no `--mem` flag: memory comes from the number of
cores you ask for.

It took about **21 minutes** to run when we did it, most of that spent reading
the fragment file, which it does several times. Ask for more time than you
expect to need — a job that hits its limit is killed part way through.

When R runs out of memory on the cluster, the scheduler kills it with no error
message — the session simply stops mid-command. If that ever happens to you,
here or anywhere else, memory is the first thing to check.

## Setup

```{r}
#| label: setup
#| cache: false
library(Signac)
library(Seurat)
library(EnsDb.Hsapiens.v86)
# GenomeInfoDb provides standardChromosomes(), seqlevels() and genome().
# Signac imports these internally but does not attach them for you.
library(GenomeInfoDb)
library(GenomicRanges)
library(ggplot2)
library(patchwork)
library(dplyr)

set.seed(1234)

# CHANGE THIS to your NetID.
NETID <- "your_netid"


WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "signac")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

SHARED <- "/groups/darrenc/cmm_523/references/pbmc_multiome"

WORK
```

## The data

In this tutorial we will use scATAC-seq data generated on about 10,000 PBMCs
with a 10x Genomics method (we'll explain it better later in the course).

Three files:

- the **feature-by-cell matrix**, containing both gene expression and peaks
- the **ATAC fragment file**, a record of every sequenced fragment and the cell
  it came from
- its **index**, which lets Signac read regions of the fragment file without
  loading all of it

The fragment file is the one with no RNA equivalent, and it is what makes the
coverage plots at the end of this tutorial possible. The peak matrix has already
thrown away everything outside a peak; the fragment file still has it.

```{r}
#| label: data
files <- c(
  "pbmc_granulocyte_sorted_10k_filtered_feature_bc_matrix.h5",
  "pbmc_granulocyte_sorted_10k_atac_fragments.tsv.gz",
  "pbmc_granulocyte_sorted_10k_atac_fragments.tsv.gz.tbi"
)

base_url <- paste0("https://cf.10xgenomics.com/samples/cell-arc/1.0.0/",
                   "pbmc_granulocyte_sorted_10k/")

get_file <- function(f) {
  shared <- file.path(SHARED, f)
  if (file.exists(shared)) return(shared)

  local <- file.path(WORK, f)
  if (!file.exists(local)) {
    message("Downloading ", f, " ...")
    download.file(paste0(base_url, f), local, mode = "wb")
  }
  local
}

counts_file <- get_file(files[1])
frag_file   <- get_file(files[2])
invisible(get_file(files[3]))   # index must sit beside the fragment file
```

```{r}
#| label: load
inputdata <- Read10X_h5(counts_file)
names(inputdata)

atac_counts <- inputdata$Peaks
dim(atac_counts)
atac_counts[1:5, 1:3]
```

Look at those values. The dots are zeros, so the preview we can see is all
zeros! That is what sparse, near-binary data means in practice, and it is why the
RNA toolkit does not transfer.

## Gene annotation

Signac needs to know where genes are so it can label plots and compute gene
activity. We pull that from an Ensembl database package.

```{r}
#| label: annotation
# This dataset was mapped against GRCh38-2020-A, so we need an hg38 annotation.
# EnsDb.Hsapiens.v86 is hg38; EnsDb.Hsapiens.v75 is hg19. Getting this wrong is
# not a loud failure -- the code runs, TSS enrichment is computed against
# coordinates that do not correspond to your peaks, every cell scores near zero,
# and your QC filter then silently discards the entire dataset. Always check
# which build your data was aligned to before choosing an annotation.
annotation <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)

# Ensembl calls chromosomes "1", "2", "X". UCSC's hg38 calls them "chr1",
# "chr2", "chrX". Rename directly rather than using seqlevelsStyle() -- the
# style-conversion route depends on GenomeInfoDb's mapping tables and is a
# recurring source of the error "Annotation genome does not match genome of
# the object".
seqlevels(annotation) <- paste0("chr", seqlevels(annotation))
genome(annotation) <- "hg38"

annotation[1:3]
```

## Building the object

```{r}
#| label: chromatin-assay
# Keep only the standard chromosomes. Peaks on scaffolds and alt contigs are
# not useful here and cause problems downstream.
grange_counts <- StringToGRanges(rownames(atac_counts), sep = c(":", "-"))
grange_use    <- seqnames(grange_counts) %in% standardChromosomes(grange_counts)
atac_counts   <- atac_counts[as.vector(grange_use), ]

chrom_assay <- CreateChromatinAssay(
  counts = atac_counts,
  sep = c(":", "-"),
  genome = "hg38",
  fragments = frag_file,
  min.cells = 10,
  min.features = 200,
  annotation = annotation
)

pbmc <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "peaks"
)

pbmc
```

```{r}
#| label: granges
granges(pbmc)[1:5]
```

Each feature is a genomic interval rather than a gene name. That is the other
structural difference from RNA data, and it is why you can ask questions of ATAC
data — what is near this peak, do these two peaks overlap — that make no sense
for a gene count matrix.

## Quality control

ATAC has its own QC metrics, and they measure things that have no RNA analogue.

```{r}
#| label: qc-metrics
# fast = FALSE keeps the per-position signal around the TSS, not just the
# summary score. TSSPlot() needs it; the default (fast = TRUE) computes the
# score and throws the profile away, and TSSPlot() then fails with
# "Position enrichment matrix not present in assay".
pbmc <- TSSEnrichment(object = pbmc, fast = FALSE)

head(pbmc@meta.data[, c("nCount_peaks", "TSS.enrichment")], 5)
```

Note where this came from: it was computed from the fragment file, not read out
of a metadata table. Cell Ranger ARC does not ship the per-barcode QC csv that
the older ATAC pipeline did, and relying on a file's columns is fragile anyway —
a renamed column gives you `NA`, and `NA < 0.05` quietly removes every cell in
your dataset.

**TSS enrichment** is the ratio of signal at transcription start sites to
flanking background. Promoters are open in essentially every cell, so this works
as close to a universal positive control: low TSS enrichment means the assay did
not work in that cell, however many reads it has.

```{r}
#| label: qc-tss
pbmc$high.tss <- ifelse(pbmc$TSS.enrichment > 2, "High", "Low")
TSSPlot(pbmc, group.by = "high.tss") + NoLegend()
```

This is the plot worth dwelling on. Each panel shows the average accessibility
profile across all transcription start sites, centred on the TSS. In the
high-scoring cells there is a sharp peak right at the TSS — the assay found the
promoters, as it should. In the low-scoring cells that peak is flattened: the
signal has degraded towards background, and those cells are telling you much
less about where the chromatin is open. You can see the failure directly here,
rather than inferring it from a single summary number.

You will also see other ATAC QC metrics in the literature — nucleosome signal,
for instance, which measures the fragment length pattern. We are not using it
here. In practice it shifts only subtly between good and bad cells, and TSS
enrichment separates them more clearly.

```{r}
#| label: qc-violin
#| fig-width: 10
#| fig-height: 5
VlnPlot(
  object = pbmc,
  features = c("nCount_peaks", "TSS.enrichment"),
  pt.size = 0.1,
  ncol = 2
)
```

Look at the distributions before choosing thresholds. A cut is only sensible
relative to what it is being applied to.

```{r}
#| label: qc-distributions
qc <- pbmc@meta.data[, c("nCount_peaks", "TSS.enrichment")]
summary(qc)
cat("cells with any NA in a QC metric:", sum(!complete.cases(qc)), "\n")
```

```{r}
#| label: qc-filter
before <- ncol(pbmc)

keep <- pbmc$nCount_peaks   > 1000 &
        pbmc$nCount_peaks   < 100000 &
        pbmc$TSS.enrichment > 1

# What each criterion removes on its own. If one is discarding almost
# everything, that is the threshold to question -- or the sign that the metric
# was not computed the way you assumed.
data.frame(
  criterion = c("counts > 1000", "counts < 100000", "TSS > 1"),
  n_passing = c(
    sum(pbmc$nCount_peaks   > 1000,   na.rm = TRUE),
    sum(pbmc$nCount_peaks   < 100000, na.rm = TRUE),
    sum(pbmc$TSS.enrichment > 1,      na.rm = TRUE)
  ),
  of_total = before
)

pbmc <- pbmc[, which(keep)]
cat("kept", ncol(pbmc), "of", before, "cells\n")

# Fail here rather than several steps downstream. A filter that removes almost
# everything is a mistake, not a result.
stopifnot(ncol(pbmc) > 500)

pbmc
```

## Normalization and dimensionality reduction

Here is where ATAC diverges most sharply from RNA.

```{r}
#| label: lsi
pbmc <- RunTFIDF(pbmc)
pbmc <- FindTopFeatures(pbmc, min.cutoff = "q0")
pbmc <- RunSVD(pbmc)

# How many components did we actually get? RunSVD() asks for 50 by default but
# can return fewer if something is wrong, and asking for dimensions that do not
# exist fails later with "subscript out of bounds" -- an error that points at
# RunUMAP rather than at the real cause here.
n_lsi <- ncol(Embeddings(pbmc, "lsi"))
n_lsi
```

**TF-IDF** comes from document retrieval. Treat each cell as a document and each
peak as a word: a peak open in every cell is uninformative, like the word "the",
while a peak open in a few cells is highly informative. TF-IDF weights peaks by
how rare they are. This is a better fit than log-normalization, which assumes a
count distribution ATAC data does not have.

**SVD on the TF-IDF matrix** is latent semantic indexing — the same technique
used for finding documents by topic. Together, TF-IDF followed by SVD is what
people mean by LSI in the ATAC literature.

```{r}
#| label: depth-correlation
DepthCor(pbmc)
```

I wanted to show you this plot because of a specific, well-known artefact:
**the first LSI component captures sequencing depth rather than biology.** This
is because we do not center the data in LSI. It is probably best practice to
always check, but we always ignore the first component for this reason, which is
why everything below uses `dims = 2:30` rather than `1:30`.

Nothing warns you about this. Include component 1 and your clusters will
partly reflect how deeply each cell was sequenced, and the UMAP will look
perfectly reasonable.

Everything below therefore starts at component 2.

```{r}
#| label: cluster
# Skip component 1 (depth), use what is available up to 30.
use_dims <- 2:min(30, n_lsi)
range(use_dims)

pbmc <- RunUMAP(object = pbmc, reduction = "lsi", dims = use_dims)
pbmc <- FindNeighbors(object = pbmc, reduction = "lsi", dims = use_dims)
pbmc <- FindClusters(object = pbmc, algorithm = 3, resolution = 1.2, verbose = FALSE)

DimPlot(object = pbmc, label = TRUE) + NoLegend()
```

## Gene activity

Clusters of peaks are hard to interpret directly. The usual move is to summarize
accessibility over each gene body and promoter, producing a matrix that looks
like gene expression and can be handled with familiar tools.

```{r}
#| label: gene-activity
gene.activities <- GeneActivity(pbmc)

pbmc[["RNA"]] <- CreateAssayObject(counts = gene.activities)
pbmc <- NormalizeData(
  object = pbmc,
  assay = "RNA",
  normalization.method = "LogNormalize",
  scale.factor = median(pbmc$nCount_RNA)
)

dim(gene.activities)
```

I want to be clear about what this is: a **proxy**, not a measurement.
Accessibility over a gene body correlates with expression, loosely. A gene can be
open and not transcribed. Gene activity is useful for recognizing cell types; it
is not a direct substitute for measuring RNA.

```{r}
#| label: gene-activity-plot
#| fig-height: 8
DefaultAssay(pbmc) <- "RNA"

FeaturePlot(
  object = pbmc,
  features = c("MS4A1", "CD3D", "LEF1", "NKG7", "TREM1", "LYZ"),
  pt.size = 0.1,
  max.cutoff = "q95",
  ncol = 3
)
```

These are known markers for expected cell types: `MS4A1` for B cells, `CD3D` for
T cells, `LYZ` for monocytes. The
signal is noisier than the RNA equivalent, which is what you should expect from
a proxy.

## Coverage plots

This is the visualization with no RNA counterpart, and it is the reason the
fragment file was worth keeping.

```{r}
#| label: coverage-cd3
DefaultAssay(pbmc) <- "peaks"

CoveragePlot(
  object = pbmc,
  region = "CD3D",
  extend.upstream = 5000,
  extend.downstream = 5000
)
```

Each row is a cluster, and the height is accessibility across the locus. You are
looking at the regulatory landscape of a gene, cell type by cell type: which
promoter is open, which enhancers are used, and by whom. A `FeaturePlot` tells
you a gene is on in T cells. This tells you *where* the chromatin is open to
make that happen.

Here is the same view for a B cell marker:

```{r}
#| label: coverage
CoveragePlot(
  object = pbmc,
  region = "MS4A1",
  extend.upstream = 5000,
  extend.downstream = 5000
)
```

## Save your work

```{r}
#| label: save
saveRDS(pbmc, file = file.path(WORK, "output", "pbmc_atac.rds"))
```

Keep this. We'll use it in a later tutorial.

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the [Signac PBMC](https://stuartlab.org/signac/articles/pbmc_vignette)
and [multiomic](https://stuartlab.org/signac/articles/pbmc_multiomic) vignettes,
updated for Signac 1.17 and Seurat 5.*
