---
title: "Joint RNA and ATAC analysis"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  # Caching is OFF. knitr stores cached objects in a lazy-load database that
  # cannot hold a variable over 2 GB, and a multiome object with ~100,000 peaks
  # exceeds that. See the note in the Signac tutorial.
  cache: false
knitr:
  opts_chunk:
    fig.path: "figs/08_rna_atac-"
    comment: "#>"
---

The [Signac tutorial](07_signac.md) analyzed chromatin accessibility on its own,
and the [Seurat intro](01_seurat_intro.md) analyzed expression on its own. This
one uses both, measured **in the same cells**.

That last part is what makes multiome data different from having an RNA dataset
and an ATAC dataset. With separate experiments you can integrate them — match
cells that look similar across modalities and hope the matching is right. With
multiome you do not have to: cell barcode `AAACAGCCAAGGAATC-1` has an expression
profile and an accessibility profile, and they came from the same nucleus.

That removes a whole category of uncertainty, and it lets you ask questions that
are otherwise out of reach. Chief among them: **which regulatory elements control
which genes?** Answering that requires knowing, cell by cell, whether a peak
being open coincides with a gene being expressed. You cannot do that if your
peaks and your genes came from different cells.

## Before you start

This tutorial needs about **48 GB of memory** and runs in **60–90 minutes**.

```bash
interactive -a cusanovichlab -n 12 -t 03:00:00
```

`interactive` allocates memory per core, so `-n 12` at the default 4 GB per core
gives you 48 GB.

## Setup

```{r}
#| label: setup
#| cache: false
library(Signac)
library(Seurat)
library(EnsDb.Hsapiens.v86)
library(GenomeInfoDb)
library(GenomicRanges)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggplot2)
library(patchwork)
library(dplyr)

set.seed(1234)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

if (nzchar(Sys.getenv("CMM523_NETID"))) NETID <- Sys.getenv("CMM523_NETID")

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "rna_atac")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

SHARED <- "/groups/darrenc/cmm_523/references/pbmc_multiome"

WORK
```

## Loading both modalities

Same dataset as the Signac tutorial, but this time we keep the gene expression
as well as the peaks.

```{r}
#| label: data
h5_file   <- file.path(SHARED, "pbmc_granulocyte_sorted_10k_filtered_feature_bc_matrix.h5")
frag_file <- file.path(SHARED, "pbmc_granulocyte_sorted_10k_atac_fragments.tsv.gz")

stopifnot(file.exists(h5_file), file.exists(frag_file))

inputdata <- Read10X_h5(h5_file)
names(inputdata)

rna_counts  <- inputdata$`Gene Expression`
atac_counts <- inputdata$Peaks

dim(rna_counts)
dim(atac_counts)
```

Two matrices, same columns. Confirm that rather than assume it — this is the
property the entire tutorial depends on:

```{r}
#| label: check-barcodes
identical(colnames(rna_counts), colnames(atac_counts))
```

## Building the object

```{r}
#| label: build
pbmc <- CreateSeuratObject(counts = rna_counts, assay = "RNA")
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

annotation <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevels(annotation) <- paste0("chr", seqlevels(annotation))
genome(annotation) <- "hg38"

grange_counts <- StringToGRanges(rownames(atac_counts), sep = c(":", "-"))
grange_use    <- seqnames(grange_counts) %in% standardChromosomes(grange_counts)
atac_counts   <- atac_counts[as.vector(grange_use), ]

pbmc[["ATAC"]] <- CreateChromatinAssay(
  counts = atac_counts,
  sep = c(":", "-"),
  genome = "hg38",
  fragments = frag_file,
  min.cells = 10,
  annotation = annotation
)

pbmc
```

One object, two assays. `DefaultAssay()` decides which one a function acts on,
and forgetting to set it is the commonest mistake in multimodal work — the code
runs, it just answers about the wrong modality.

## Quality control on both modalities

A cell has to pass QC in both assays to be usable. Failing either one makes it
useless for joint analysis, however good the other looks.

```{r}
#| label: qc-metrics
DefaultAssay(pbmc) <- "ATAC"
pbmc <- NucleosomeSignal(pbmc)
pbmc <- TSSEnrichment(pbmc, fast = FALSE)

summary(pbmc@meta.data[, c("nCount_RNA", "nCount_ATAC", "percent.mt",
                           "TSS.enrichment", "nucleosome_signal")])
```

```{r}
#| label: qc-violin
#| fig-width: 14
#| fig-height: 5
VlnPlot(
  pbmc,
  features = c("nCount_RNA", "nCount_ATAC", "percent.mt",
               "TSS.enrichment", "nucleosome_signal"),
  ncol = 5, log = TRUE, pt.size = 0
) + NoLegend()
```

```{r}
#| label: qc-filter
before <- ncol(pbmc)

keep <- pbmc$nCount_ATAC       < 1e5 &
        pbmc$nCount_ATAC       > 1000 &
        pbmc$nCount_RNA        < 25000 &
        pbmc$nCount_RNA        > 1000 &
        pbmc$percent.mt        < 20 &
        pbmc$nucleosome_signal < 2 &
        pbmc$TSS.enrichment    > 1

data.frame(
  criterion = c("ATAC < 1e5", "ATAC > 1000", "RNA < 25000", "RNA > 1000",
                "percent.mt < 20", "nucleosome < 2", "TSS > 1"),
  n_passing = c(
    sum(pbmc$nCount_ATAC       < 1e5,   na.rm = TRUE),
    sum(pbmc$nCount_ATAC       > 1000,  na.rm = TRUE),
    sum(pbmc$nCount_RNA        < 25000, na.rm = TRUE),
    sum(pbmc$nCount_RNA        > 1000,  na.rm = TRUE),
    sum(pbmc$percent.mt        < 20,    na.rm = TRUE),
    sum(pbmc$nucleosome_signal < 2,     na.rm = TRUE),
    sum(pbmc$TSS.enrichment    > 1,     na.rm = TRUE)
  ),
  of_total = before
)

pbmc <- pbmc[, which(keep)]
cat("kept", ncol(pbmc), "of", before, "cells\n")
stopifnot(ncol(pbmc) > 500)
```

Note that requiring both modalities to pass costs you cells. That is the price
of multiome data, and it is worth knowing before you design an experiment around
it: you will recover fewer usable cells than you would from either assay run
alone.

## Processing each modality

Each gets the treatment appropriate to its data type — SCTransform and PCA for
RNA, TF-IDF and LSI for ATAC. Same object, different pipelines.

```{r}
#| label: process-rna
DefaultAssay(pbmc) <- "RNA"
pbmc <- SCTransform(pbmc, verbose = FALSE)
pbmc <- RunPCA(pbmc)
pbmc <- RunUMAP(pbmc, dims = 1:50, reduction.name = "umap.rna",
                reduction.key = "rnaUMAP_")
```

```{r}
#| label: process-atac
DefaultAssay(pbmc) <- "ATAC"
pbmc <- RunTFIDF(pbmc)
pbmc <- FindTopFeatures(pbmc, min.cutoff = "q0")
pbmc <- RunSVD(pbmc)

n_lsi <- ncol(Embeddings(pbmc, "lsi"))
use_dims <- 2:min(50, n_lsi)   # component 1 tracks sequencing depth

pbmc <- RunUMAP(pbmc, reduction = "lsi", dims = use_dims,
                reduction.name = "umap.atac", reduction.key = "atacUMAP_")
```

```{r}
#| label: compare-modalities
#| fig-width: 12
p1 <- DimPlot(pbmc, reduction = "umap.rna") + ggtitle("RNA") + NoLegend()
p2 <- DimPlot(pbmc, reduction = "umap.atac") + ggtitle("ATAC") + NoLegend()
p1 + p2
```

Two views of the same cells. They agree on the broad structure and disagree on
the details, which is the interesting part — the modalities are measuring
related but distinct things, and where they disagree is often where the biology
is.

## Annotating cell types

```{r}
#| label: annotate
DefaultAssay(pbmc) <- "SCT"
pbmc <- FindNeighbors(pbmc, dims = 1:50)
pbmc <- FindClusters(pbmc, resolution = 0.8, verbose = FALSE)

DimPlot(pbmc, reduction = "umap.rna", label = TRUE) + NoLegend()
```

```{r}
#| label: markers
#| fig-height: 10
FeaturePlot(
  pbmc, reduction = "umap.rna",
  features = c("MS4A1", "CD3D", "CD8A", "NKG7", "LYZ", "FCGR3A",
               "CD14", "FCER1A", "PPBP"),
  ncol = 3
)
```

## Linking peaks to genes

This is the analysis multiome data exists for. For each gene, `LinkPeaks()`
looks at nearby peaks and asks whether accessibility correlates with expression
across cells — while controlling for GC content, overall accessibility, and peak
size, so that the result is not simply driven by which peaks are big or open in
everything.

```{r}
#| label: link-peaks
DefaultAssay(pbmc) <- "ATAC"

pbmc <- RegionStats(pbmc, genome = BSgenome.Hsapiens.UCSC.hg38)

pbmc <- LinkPeaks(
  object = pbmc,
  peak.assay = "ATAC",
  expression.assay = "SCT",
  genes.use = c("MS4A1", "CD3D", "LYZ", "NKG7")
)

links <- Links(pbmc)
length(links)
head(as.data.frame(links), 10)
```

Restricting to four genes keeps this fast. Running genome-wide is the same call
without `genes.use`, and it takes hours rather than minutes.

```{r}
#| label: coverage-links
#| fig-height: 8
CoveragePlot(
  object = pbmc,
  region = "MS4A1",
  features = "MS4A1",
  expression.assay = "SCT",
  extend.upstream = 10000,
  extend.downstream = 5000
)
```

Read this plot carefully, because it is the payoff of the whole tutorial. The
coverage tracks show accessibility per cell type. The panel on the right shows
expression of the same gene in the same cells. The arcs at the bottom are the
inferred peak-to-gene links. You are looking at a regulatory hypothesis: *this*
element, open in *these* cells, appears to drive *this* gene.

```{r}
#| label: coverage-cd3d
#| fig-height: 8
CoveragePlot(
  object = pbmc,
  region = "CD3D",
  features = "CD3D",
  expression.assay = "SCT",
  extend.upstream = 10000,
  extend.downstream = 5000
)
```

## What a link is and is not

A link is a correlation across cells, computed with a null model that accounts
for some obvious confounders. It is not evidence that the element regulates the
gene.

Two peaks near each other tend to be open in the same cells whether or not
either controls the gene. A peak can correlate with expression because both
respond to the same upstream signal. And the method only considers peaks within
a fixed window, so genuine long-range interactions are invisible to it by
construction.

What links are good for is narrowing the field. Genome-wide you might have
100,000 peaks and no idea which matter for your gene of interest. Linking gives
you five candidates worth testing — and testing means a perturbation experiment,
not a smaller p-value.

## Save your work

```{r}
#| label: save
saveRDS(pbmc, file = file.path(WORK, "output", "pbmc_multiome.rds"))
```

The [WNN tutorial](09_wnn.md) starts from this object.

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the [Signac multiomic
vignette](https://stuartlab.org/signac/articles/pbmc_multiomic), updated for
Signac 1.17 and Seurat 5.*
