---
title: "Weighted nearest neighbors and motif activity"
subtitle: "CMM 523 — Crash Course in Single-cell Genomics Analysis"
format:
  gfm:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  # Caching is OFF. knitr's lazy-load cache cannot hold a variable over 2 GB,
  # and a multiome object with ~100,000 peaks exceeds that.
  cache: false
knitr:
  opts_chunk:
    fig.path: "figs/09_wnn-"
    comment: "#>"
---

The [previous tutorial](08_rna_atac.md) processed RNA and ATAC separately and
produced two UMAPs of the same cells. That leaves an awkward question: which one
do you cluster on?

Neither answer is satisfying. Cluster on RNA and you ignore the chromatin
evidence. Cluster on ATAC and you ignore expression. Concatenating the two is
worse than it sounds — the modalities have different scales, different noise
characteristics, and different numbers of features, so whichever has more
variance simply wins.

Weighted nearest neighbors solves this per cell rather than globally. For each
individual cell, WNN asks which modality better predicts its neighbors, and
weights accordingly. A cell whose identity is clear from expression leans on RNA;
one distinguished mainly by its regulatory state leans on ATAC. The weights are
learned from the data, not set by you.

This tutorial then does something ATAC makes possible and RNA does not: infers
which **transcription factors** are active in each cell, by asking which binding
motifs are enriched in the regions that cell has open.

## Before you start

This tutorial is the most demanding in the course. It needs about **64 GB of
memory** and runs in **3–5 hours**, most of that in the chromVAR step.

```bash
interactive -a cusanovichlab -n 16 -t 08:00:00
```

Do not run this interactively and wait. Write a batch script and come back.

## Setup

```{r}
#| label: setup
#| cache: false
library(Signac)
library(Seurat)
library(EnsDb.Hsapiens.v86)
library(GenomeInfoDb)
library(GenomicRanges)
library(BSgenome.Hsapiens.UCSC.hg38)
library(chromVAR)
library(SummarizedExperiment)
library(BiocParallel)
library(motifmatchr)
library(TFBSTools)
library(JASPAR2024)
library(RSQLite)
library(presto)
library(ggplot2)
library(patchwork)
library(dplyr)

set.seed(1234)

# CHANGE THIS to your NetID.
NETID <- "your_netid"

if (nzchar(Sys.getenv("CMM523_NETID"))) NETID <- Sys.getenv("CMM523_NETID")

WORK <- file.path("/xdisk/darrenc/cmm_523", NETID, "wnn")
dir.create(file.path(WORK, "output"), recursive = TRUE, showWarnings = FALSE)

PREV <- file.path("/xdisk/darrenc/cmm_523", NETID, "rna_atac", "output",
                  "pbmc_multiome.rds")

options(future.globals.maxSize = 16000 * 1024^2)

# Run chromVAR serially. It parallelizes through BiocParallel, and each worker
# forks a copy of the peak matrix -- with ~100,000 peaks and 11,000 cells that
# multiplies memory fast and gets the job killed by the scheduler. The failure
# does not look like a memory error either: you get
# "wrong args for environment subassignment" from BiocParallel's result
# collector, because it is trying to gather a result from a worker that no
# longer exists. Serial is slower and it finishes.
BiocParallel::register(BiocParallel::SerialParam())

WORK
```

## Starting from the previous tutorial

```{r}
#| label: load
stopifnot(file.exists(PREV))
pbmc <- readRDS(PREV)
pbmc
```

If that `stopifnot` fails, run the [RNA+ATAC tutorial](08_rna_atac.md) first —
this one starts where that one stopped.

## Weighted nearest neighbors

```{r}
#| label: wnn
pbmc <- FindMultiModalNeighbors(
  pbmc,
  reduction.list = list("pca", "lsi"),
  dims.list = list(1:50, 2:50)
)

pbmc <- RunUMAP(pbmc, nn.name = "weighted.nn",
                reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")

pbmc <- FindClusters(pbmc, graph.name = "wsnn", algorithm = 3, verbose = FALSE)
```

Note `dims.list = list(1:50, 2:50)`. The RNA reduction uses all 50 components;
the ATAC one starts at 2, because the first LSI component tracks sequencing
depth rather than biology. That asymmetry is easy to miss and it matters.

```{r}
#| label: three-umaps
#| fig-width: 16
#| fig-height: 5
p1 <- DimPlot(pbmc, reduction = "umap.rna",  label = TRUE, label.size = 3) +
  ggtitle("RNA") + NoLegend()
p2 <- DimPlot(pbmc, reduction = "umap.atac", label = TRUE, label.size = 3) +
  ggtitle("ATAC") + NoLegend()
p3 <- DimPlot(pbmc, reduction = "wnn.umap",  label = TRUE, label.size = 3) +
  ggtitle("WNN") + NoLegend()

p1 + p2 + p3
```

## Looking at the weights

The weights are the interesting output, and they are easy to overlook.

`FindMultiModalNeighbors()` writes one weight column per modality into the
metadata, named after the **assay** behind each reduction — not after the
reduction. Our PCA was computed on the `SCT` assay, so the column is
`SCT.weight` rather than `RNA.weight`. Find it rather than assuming:

```{r}
#| label: weight-columns
weight_cols <- grep("\\.weight$", colnames(pbmc@meta.data), value = TRUE)
weight_cols

rna_weight <- weight_cols[1]
summary(pbmc@meta.data[[rna_weight]])
```

```{r}
#| label: weights
#| fig-height: 6
VlnPlot(pbmc, features = rna_weight, group.by = "seurat_clusters",
        sort = TRUE, pt.size = 0.1) +
  NoLegend() +
  labs(title = paste(rna_weight, "by cluster"),
       subtitle = "1 = decided entirely by expression, 0 = entirely by chromatin")
```

```{r}
#| label: weight-umap
FeaturePlot(pbmc, reduction = "wnn.umap", features = rna_weight) +
  scale_colour_viridis_c() +
  ggtitle(paste("Per-cell", rna_weight))
```

Cells on the right of that violin plot were defined mostly by expression; those
on the left by chromatin. This is a real result about your data rather than a
diagnostic. If a population sits near 0.5, both modalities contributed — and if
one sits at an extreme, that tells you something about which measurement carries
the information for that cell type.

## Motif analysis

Everything so far treated peaks as anonymous coordinates. Now we ask what is
*in* them.

The logic: transcription factors bind specific sequence motifs. If a cell's open
regions are enriched for a particular motif, that factor is plausibly active in
that cell. This gives a per-cell estimate of TF activity — something RNA alone
cannot provide, because a transcription factor's mRNA level is a poor guide to
whether the protein is bound and working.

```{r}
#| label: jaspar
DefaultAssay(pbmc) <- "ATAC"

# JASPAR2024 changed its interface. getMatrixSet() will NOT accept a JASPAR2024
# object directly -- it fails with "unable to find an inherited method for
# function 'getMatrixSet' for signature 'JASPAR2024'". You have to open a
# connection to the bundled SQLite database and pass that instead. Older code
# and tutorials using JASPAR2020 pass the object, which is why this is a common
# thing to get stuck on.
jaspar     <- JASPAR2024::JASPAR2024()
jaspar_con <- RSQLite::dbConnect(RSQLite::SQLite(), db(jaspar))

pwm_set <- TFBSTools::getMatrixSet(
  jaspar_con,
  opts = list(species = 9606, collection = "CORE", all_versions = FALSE)
)

length(pwm_set)
```

```{r}
#| label: add-motifs
motif.matrix <- CreateMotifMatrix(
  features = granges(pbmc),
  pwm = pwm_set,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  use.counts = FALSE
)

motif.object <- CreateMotifObject(data = motif.matrix, pwm = pwm_set)
Motifs(pbmc[["ATAC"]]) <- motif.object

dim(motif.matrix)
```

That matrix is peaks by motifs: for each peak, which transcription factor motifs
it contains.

Older tutorials do this with `Signac::RunChromVAR()`. That wrapper was removed
in Signac 1.17, so we call chromVAR directly. It is no harder, and it makes
visible what the wrapper was hiding — which is worth seeing once.

```{r}
#| label: chromvar
# This is the slow step: 30-60 minutes.

# 1. chromVAR works on a SummarizedExperiment, not a Seurat object. Build one
#    from the peak counts and their genomic ranges.
peak_counts <- LayerData(pbmc, assay = "ATAC", layer = "counts")
peak_ranges <- granges(pbmc)

se <- SummarizedExperiment(
  assays = list(counts = peak_counts),
  rowRanges = peak_ranges
)

# 2. Compute GC content per peak. This is the correction that stops the result
#    being a readout of sequence composition -- GC-rich regions are more
#    accessible in essentially every assay, so without this the motifs that
#    happen to be GC-rich would look universally "active".
se <- addGCBias(se, genome = BSgenome.Hsapiens.UCSC.hg38)

# 3. The peak-by-motif matrix we built above, pulled back out of the object.
motif_ix <- GetMotifData(object = pbmc, assay = "ATAC", slot = "data")

# 4. Deviations. chromVAR compares observed accessibility at each motif's sites
#    against a background set of peaks matched for GC content and overall
#    accessibility, then reports a z-score per motif per cell.
dev <- computeDeviations(object = se, annotations = motif_ix)

# That took an hour, so write it out. Saving any result that was expensive to
# compute is a good habit -- you will want it again.
saveRDS(dev, file.path(WORK, "output", "chromvar_deviations.rds"))

# 5. Put the z-scores into the Seurat object as their own assay.
pbmc[["chromvar"]] <- CreateAssayObject(data = assays(dev)$z)

DefaultAssay(pbmc) <- "chromvar"
pbmc
```

So the deviation score is: how much more (or less) accessible that motif's sites
are in this cell than in comparable peaks. Those background corrections are the
whole method — without them you would mostly be measuring which cells have more
open chromatin overall, and every motif would look active in the same cells.

## TF expression versus motif activity

Here is where having both modalities pays off, because you can compare a
factor's mRNA against its inferred activity in the same cells.

```{r}
#| label: motif-vs-expression
#| fig-width: 12
# assay = "ATAC" is not optional here. The motif data lives in the ATAC assay,
# but the default assay is now "chromvar" -- and ConvertMotifID looks in the
# default unless told otherwise, failing with "Cannot run ConvertMotifID on a
# standard Assay object". Anywhere you touch motifs after this point, name the
# assay.
motif.name <- ConvertMotifID(pbmc, name = "CEBPB", assay = "ATAC")

DefaultAssay(pbmc) <- "SCT"
gene_plot <- FeaturePlot(pbmc, features = "CEBPB", reduction = "wnn.umap") +
  ggtitle("CEBPB expression (RNA)")

DefaultAssay(pbmc) <- "chromvar"
motif_plot <- FeaturePlot(pbmc, features = motif.name, reduction = "wnn.umap",
                          min.cutoff = 0, cols = c("lightgrey", "darkred")) +
  ggtitle("CEBPB motif activity (ATAC)")

gene_plot | motif_plot
```

These two panels do not have to agree, and where they disagree is informative. A
factor can be transcribed without being active — held inactive by phosphorylation
state, localization, or a missing cofactor. It can also be active with low mRNA,
because a little protein goes a long way and mRNA is noisy in single cells.

## Finding cell-type-specific factors

The strongest evidence for a factor mattering in a cell type is when expression
and motif activity agree. We test both and intersect.

```{r}
#| label: markers-both
markers_rna <- presto::wilcoxauc(pbmc, group_by = "seurat_clusters",
                                 seurat_assay = "SCT", assay = "data")
markers_motifs <- presto::wilcoxauc(pbmc, group_by = "seurat_clusters",
                                    seurat_assay = "chromvar", assay = "data")

motif.names <- markers_motifs$feature
colnames(markers_rna)    <- paste0("RNA.",   colnames(markers_rna))
colnames(markers_motifs) <- paste0("motif.", colnames(markers_motifs))
markers_rna$gene    <- markers_rna$RNA.feature
markers_motifs$gene <- ConvertMotifID(pbmc, id = motif.names, assay = "ATAC")

topTFs <- function(celltype, padj.cutoff = 1e-2) {
  ctmarkers_rna <- dplyr::filter(
    markers_rna, RNA.group == celltype,
    RNA.padj < padj.cutoff, RNA.logFC > 0) %>% arrange(-RNA.auc)

  ctmarkers_motif <- dplyr::filter(
    markers_motifs, motif.group == celltype,
    motif.padj < padj.cutoff, motif.logFC > 0) %>% arrange(-motif.auc)

  top_tfs <- inner_join(
    x = ctmarkers_rna[, c("RNA.feature", "gene", "RNA.auc", "RNA.padj")],
    y = ctmarkers_motif[, c("motif.feature", "gene", "motif.auc", "motif.padj")],
    by = "gene"
  )
  top_tfs$avg_auc <- (top_tfs$RNA.auc + top_tfs$motif.auc) / 2
  arrange(top_tfs, -avg_auc)
}

for (ct in head(levels(pbmc$seurat_clusters), 6)) {
  cat("\n--- cluster", ct, "---\n")
  print(head(topTFs(ct)[, c("gene", "RNA.auc", "motif.auc", "avg_auc")], 4))
}
```

A factor appearing here is expressed *and* has enriched motif accessibility in
that cluster. That is much better evidence than either alone — though still
correlative, and still a hypothesis rather than a result.

## What motif activity cannot tell you

Worth being explicit, because chromVAR output is easy to over-read.

Motifs are short and degenerate. A given motif occurs in far more places in the
genome than the factor actually binds, so motif presence is a weak proxy for
occupancy.

Related factors share motifs. Many JASPAR entries are near-identical across a
family, and chromVAR cannot distinguish family members. A "CEBPB" signal may
equally be CEBPA, CEBPD, or several others.

And an enriched motif does not establish direction. A factor's sites being open
is consistent with it binding there, with it having bound previously, or with
something else keeping that chromatin accessible.

What this is good for is generating a short list of candidate regulators from an
unbiased genome-wide measurement. The next step is a perturbation, a ChIP, or a
footprinting analysis — not a stronger claim from the same data.

## Save your work

```{r}
#| label: save
saveRDS(pbmc, file = file.path(WORK, "output", "pbmc_wnn_chromvar.rds"))
RSQLite::dbDisconnect(jaspar_con)
```

## Session information

```{r}
#| label: session
sessionInfo()
```

---

*Adapted from the Seurat [WNN
vignette](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis)
and the Signac [motif analysis
vignette](https://stuartlab.org/signac/articles/motif_vignette), updated for
Seurat 5, Signac 1.17 and JASPAR2024.*
